```{r}
#| include: false
library(multvardiv)
```

# Introduction

Statistical divergences, such as the Kullback–Leibler divergence (KLD), Rényi divergence (RD), Bhattacharyya divergence (BD), and Hellinger divergence (HD), are fundamental tools in information theory, machine learning and statistics. These divergences quantify the 'distance' or dissimilarity between probability distributions, enabling a wide range of applications. 
The Kullback–Leibler divergence (KLD) is widely used in information theory, pattern recognition, and machine learning (for example, in training generative models and reinforcement learning), as well as in model comparison, variational inference, and other related fields in computer science. It is used to assess the effectiveness of probabilistic models, compare distributions, and perform tasks such as feature selection. 
The Rényi divergence (RD) is a generalization of the KL divergence. RD is used for privacy and security (differential privacy) and hypothesis testing.
The Bhattacharyya divergence (BD) is used for clustering and classification (feature selection and clustering algorithms), signal processing (e.g. signal detection in radar or sonar), and computer vision (e.g. image retrieval and object recognition, and feature distribution comparison).
The Hellinger divergence (HD) is used in robust statistics (it is less sensitive to outliers than the KL divergence), domain adaptation (transfer learning), nonparametric testing (goodness-of-fit tests and two-sample tests) and machine learning (e.g. adversarial learning).

These divergences can be computed for both univariate or multivariate distributions in either discrete or continuous cases. The discrete scenario presents a simpler calculation as it bypasses the need for integral computation.
Explicit expressions of most statistical divergences have been developed for many continuous univariate distributions, such as the gamma distribution, the Cauchy distribution, the generalized gamma distribution, the generalized Gaussian distribution, and so on. However, statistical divergences for continuous multivariate distributions require dealing with integrals whose domain is the set of all symmetric, positive definite matrices. Unfortunately, closed-form expressions are not easily found.   
Very recently, closed form expressions of the KLD have been developed
for the multivariate generalized Gaussian distribution (MGGDs) [@bouhlel19] and multivariate Cauchy distributions (MCDs) [@bouhlel22]. Furthermore, an expression
of the RD has been developed for Multivariate $t$-distributions (MTDs) [@bouhlel23]. The BD and HD can be computed from the RD. In this paper, we propose an efficient implementation of these divergences in R packages that is both precise and computationally fast,
[`multvardiv`](https://CRAN.R-project.org/package=multvardiv). Our
approach is very different from functions provided by existing R packages such as
[`philentropy`](https://joss.theoj.org/papers/10.21105/joss.00765),
[`FNN`](https://CRAN.R-project.org/package=FNN),
[`dlookr`](https://CRAN.R-project.org/package=dlookr),
[`LaplacesDemon`](https://web.archive.org/web/20150206004624/http://www.bayesian-inference.com/software),
which allow the computation of statistical divergence in discrete case [@philentropy; @FNN; @dlookr; @LaplacesDemon].  

<!-- The Kullback-Leibler divergence (KLD) is a measure of the dissimilarity -->
<!-- between two univariate or multivariate probability distributions. -->
<!-- <!-- It is extensively used on statistical signal processing, machine learning, pattern recognition and other related fields in computer science. --> 
<!-- It is extensively used on statistical signal processing, pattern -->
<!-- recognition, machine learning, data science, statistical inference and -->
<!-- other related fields in computer science, to assess the effectiveness of -->
<!-- probabilistic models, compare distributions and perform tasks like -->
<!-- feature selection. \textcolor{red}{[AJOUTER DES REF]} Explicit -->
<!-- expressions of KLD have been developed for most univariate distributions -->
<!-- such as the Gamma, the Cauchy, the generalized Gamma, the generalized -->
<!-- Gaussian ones, etc. KLD expressions have also been developed around -->
<!-- multivariate distributions such as, for example, Wishart and -->
<!-- multivariate Gaussian. -->
<!-- Other measurements of the divergencence between two probability -->
<!-- distributions include the Rényi divergence (RD), the Bhattacharyya -->
<!-- divergence (BD) and the Hellinger divergence (HD). -->

<!-- Very recently, closed form expressions of the KLD have been developed -->
<!-- for MGGDs [@bouhlel19] and MCDs [@bouhlel22]. Furthermore, an expression -->
<!-- of the RD has been developed for MTDs [@bouhlel23]. The BD and HD can be -->
<!-- computed from the RD. For scientific interest, we propose in this -->
<!-- article to implement these divergences in the R packages -->
<!-- [`multvardiv`](https://CRAN.R-project.org/package=multvardiv). Our -->
<!-- approach is very different from functions provided by existing R -->
<!-- packages such as -->
<!-- [`philentropy`](https://joss.theoj.org/papers/10.21105/joss.00765), -->
<!-- [`FNN`](https://CRAN.R-project.org/package=FNN), -->
<!-- [`dlookr`](https://CRAN.R-project.org/package=dlookr), -->
<!-- [`LaplacesDemon`](https://web.archive.org/web/20150206004624/http://www.bayesian-inference.com/software), -->
<!-- which allow the computation of KLD between discrete (or discretized) -->
<!-- probability distributions [@philentropy; @FNN; @dlookr; @LaplacesDemon]. -->


In this paper, we will focus on three multivariate distributions: MTD, MCD and MGGD.  
The MTD belongs to the elliptical symmetric distributions [@kotz; @bouhlel23]. Such distribution has been deployed in several signal and image processing applications where
heavy-tailed distribution is weel-suited.  
The MCD is a special case of the multivariate $t$-distribution [@bouhlel22]. MCD
has been recently used in several signal and image processing
applications for which non-Gaussian models are needed. These
applications include speckle denoising [@sahu; @ranjani], color image
denoising [@sadreazami16], watermarking [@sadreazami14], speech
enhancement [@fontaine].  
The MGGD belongs to the family of elliptic distributions [@pascal]. MGGD is used in several signal and image processing applications [@bouhlel19]. In particular, it
is widely used to accurately model the magnitudes of wavelet detail
coefficients to capture inter-band dependencies or/and intra-band in a
wide-sense (multi-scale, multi-channel, spatial dependencies, etc) [@Bouhlel2022]. 
Application examples include, but are not limited to: image denoising
[@sendur; @dongwook05; @dongwook09], texture image retrieving and
classification [@verdoolaege09; @verdoolaege11; @berthoumieu].

In what follows, we first present the MTD, the MCD and the MGGD
(sections [Multivariate $t$-Distribution], [Multivariate Cauchy Distribution]
and [Multivariate Generalized Gaussian Distribution]).
For each of these distributions, some results are presented,
and analytic expressions of the statistical divergences are exposed.
All of these results are used in the
[`multvardiv`](https://CRAN.R-project.org/package=multvardiv) package.  
Then we present the tools available in the
[`multvardiv`](https://CRAN.R-project.org/package=multvardiv) package
for these distributions ([The multvardiv package]).
We first present some functions for the MTDs ([Multivariate $t$-Distribution]): probability density function, simulation, parameter estimation and graphical
representations. Then we present the numerical computation of statistical
divergence between MTDs with diffrent level of precisions.
The [`multvardiv`](https://CRAN.R-project.org/package=multvardiv) package
provides the same tools for the MCD ([Multivariate Cauchy Distribution])
and  MGGD ([Multivariate generalized Gaussian distribution]).
Some particular cases provide interesting results. We use these
particular cases to test the computation of the statistical divergence
([Some particular cases -- test of the results]).

# Multivariate elliptical distributions and divergence

## Multivariate $t$-distribution

Let $\mathbf{X}$ be a random vector of $\mathbb{R}^p$ from a
multivariate $t$-distribution (MTD) with degrees of freedom $\nu$, mean
vector $\boldsymbol{\mu}$ and scale matrix $\boldsymbol{\Sigma}$, which is a
$p \times p$ symmetric, positive definite real matrix. Its probability
density function is given by: $$
    \displaystyle{ f_{\mathbf{X}}(\mathbf{x}|\nu, \boldsymbol{\mu}, \boldsymbol{\Sigma}) = \frac{\Gamma\left( \frac{\nu+p}{2} \right) |\mathbf{\Sigma}|^{-1/2}}{\Gamma\left( \frac{\nu}{2} \right) (\nu \pi)^{p/2}} \left( 1 + \frac{1}{\nu} (\mathbf{x}-\boldsymbol{\mu})^T \boldsymbol{\Sigma}^{-1} (\mathbf{x}-\boldsymbol{\mu}) \right)^{-\frac{\nu+p}{2}} }
$$ {#eq-mcd}

where $\Gamma(\cdot)$ is the gamma function.

A sample from a multivariate $t$-distribution with location parameter
$\boldsymbol{\mu}$, scale matrix $\boldsymbol{\Sigma}$ and $\nu$ degrees of freedom
can be generated using [@roth; @kotz]: $$
    \displaystyle{\mathbf{X} = \boldsymbol{\mu} + \frac{\mathbf{Y}}{\sqrt{U/\nu}}}
$$ {#eq-rmtd} where $\mathbf{Y}$ is generated from a centered Gaussian
distribution with covariance matrix $\boldsymbol{\Sigma}$, i.e. $\mathbf{Y}\sim \mathcal{N}^p(0,\mathbf{\Sigma})$ and $U$ is distributed
according to a Chi-squared distribution with $\nu$ degrees of freedom, $U\sim \chi^2(\nu)$ and $U$ is independent of $\mathbf{Y}$. For convenience, we use the following new variable: $V=\frac{U}{\nu}$. As a consequence, $V$ follows a Gamma distribution with parameters $(\frac{\nu}{2},\frac{\nu}{2})$, i.e. $V\sim Gamma(\frac{\nu}{2},\frac{\nu}{2})$. 

<!-- The MCD is a particular case of the multivariate $t$-distribution with $\nu = 1$. -->

<!-- Hence, a sample from a MCD with location parameter $\boldsymbol{\mu}$ and dispersion matrix $\Sigma$ can be generated using @eq-rmtd with $\nu = 1$. -->

### Parameter estimation by the EM method {#sec-est_mtd}

We use the Expectation-Maximization (EM) method to estimate the parameters of a multivariate $t$-distribution.
Let $\{ \mathbf{x}_1, \dots, \mathbf{x}_n \}$ be $n$ observations from an MTD with
a probability density function given by $f_{\mathbf{X}}(\mathbf{x})$. We
assume that each random vector $\mathbf{x}_i$ is associated with a non-observed, hidden variable $V_i$. The sequence $\{V_i,i=1,\cdots,N\}$ is
also assumed to be $iid$.
The probability density of $\mathbf{X}$ depends on a set of unknown
parameters $\theta = (\boldsymbol{\mu}, \boldsymbol{\Sigma}, \nu)$.
<!-- We know that $\displaystyle{\mathbf{X} = \boldsymbol{\mu} + \frac{\mathbf{Y}}{\sqrt{u/\nu}}}$ (@eq-rmtd), -->
<!-- where $\Y\sim \mathcal{N}^p(0,\gSigma)$ is a $p$-variate normal random vector with mean 0 and covariance matrix $\gSigma$, $U$ is the chi-squared -->
<!-- random variable with degrees of freedom $\nu$ and $U$ is independent of $\Y$. -->

The ML estimator $\hat{\theta}$ of the parameters $\theta$ of the
compound distribution $f_{\mathbf{X}}(\mathbf{x})$ is given by: $$
\hat{\theta}_{ML} = \arg \max_{\theta} f_{\mathbf{X}}(\mathbf{x}|\theta) = \arg \max_{\theta}\prod_{i=1}^N f_{\mathbf{X}}(\mathbf{x}_i|\theta)
$$ {#eq-MV}

In the lack of an explicit solution of @eq-MV, the EM algorithm is used
to find an estimation $\hat{\theta}$, given a current $\theta'$
estimation. The new expression is then given by: $$\begin{array}{lll}
\hat{\theta} &    = &   \arg \max_{\theta} \mathbb{E}_{V|\mathbf{X}} \left\{ \ln f_{V,\mathbf{X}}(V,\mathbf{X}|\theta)\big|\mathbf{x},\theta' \right\} \\
&   = &   \arg \max_{\theta} \mathbb{E}_{V|\mathbf{X}} \left\{ \ln \prod\limits_{i=1}^N f_{V,\mathbf{X}}(V_i,\mathbf{X}_i|\theta)\big|\mathbf{x}_i,\theta' \right\} \\
&   = &   \arg \max_{\theta} \mathbb{E}_{V|\mathbf{X}} \left\{ \sum\limits_{i=1}^N \ln f_{V,\mathbf{X}}(V_i,\mathbf{X}_i|\theta)\big|\mathbf{x}_i,\theta' \right\} \\
&   = &   \arg \max_{\theta} \sum\limits_{i=1}^N \mathbb{E}_{V|\mathbf{X}} \left\{ \ln f_{V,\mathbf{X}}(V_i, \mathbf{X}_i|\theta)\big|\mathbf{x}_i,\theta' \right\}
\end{array}$$ {#eq-armax}

Let $\theta_y=(\boldsymbol{\mu},\boldsymbol{\Sigma})$ and $\theta_v=(\nu)$, using the
fact that
$f_{V,\mathbf{X}}(v_i,\mathbf{x}_i|\theta)=f_{\mathbf{X}|V}(\mathbf{x}_i|v_i,\theta_y)f_{V}(v_i|\theta_{v})$,
then @eq-armax can bee maximised separately from the parameters
$\theta_y$ et $\theta_{v}$, so:

$$
\hat{\theta}_y = \arg \max_{\theta_y}\sum_{i=1}^N \mathbb{E}_{V|\mathbf{X}}\{\ln f_{\mathbf{X}|V}(\mathbf{X}_i|V_i,\theta_y)\big|\mathbf{x}_i,\theta'\}
$$ {#eq-lamda_w}
and
$$
\hat{\theta}_v = \arg \max_{\theta_y}\sum_{i=1}^N \mathbb{E}_{V|\mathbf{X}}\{\ln f_{V}(V_i|\theta_{v})\big|\mathbf{x}_i,\theta'\}
$$ {#eq-lamda_x}

#### Estimation of $\theta_y=(\boldsymbol{\mu},\boldsymbol{\Sigma})$ {.unnumbered .unlisted}

The probability density function of $\mathbf{X}$ knowing $v$ is given as follows 
$$\begin{array}{lll}
f_{\mathbf{X}|V}(\mathbf{x_i}|v_i, \theta_y) &    = &  \big|\frac{\partial \mathbf{Y}}{\partial \mathbf{X}}\big| f_{\mathbf{Y}}\left( (\mathbf{x_i}-\boldsymbol{\mu}) \sqrt{v_i}\right) \\
&   = &  v_i^{\frac{p}{2}}\frac{1}{(2\pi)^{p/2}|\boldsymbol{\Sigma}|^{1/2}}\exp\big(-\frac{v_i}{2}(\mathbf{x_i}-\boldsymbol{\mu})^T\boldsymbol{\Sigma}^{-1}(\mathbf{x_i}-\boldsymbol{\mu})\big)  \\
&   = & \frac{1}{(2\pi)^{p/2}|\frac{\boldsymbol{\Sigma}}{v_i}|^{1/2}}\exp\big(-\frac{v_i}{2}(\mathbf{x_i}-\boldsymbol{\mu})^T\boldsymbol{\Sigma}^{-1}(\mathbf{x_i}-\boldsymbol{\mu})\big)
\end{array}$${#eq-conditional_law}
where $\big|\frac{\partial \mathbf{Y}}{\partial \mathbf{X}}\big|$ is the Jacobian determinant.
It is clear that $\mathbf{X}|V\sim \mathcal{N}^p(\boldsymbol{\mu}, \frac{\boldsymbol{\Sigma}}{v})$. 
Substituting the expression of the probability density function of $\mathbf{X}|V$ given by @eq-conditional_law into @eq-lamda_w, differentiating with respect to $\boldsymbol{\mu}$ and $\boldsymbol{\Sigma}$, and setting the result to zero gives:
<!-- Nous allons procéder autrement, soit $V=\frac{U}{\nu}$, nous allons alors déterminer $f_{\X|V}(x_i|v_i, \theta_y)$ et $f_{V}(v_i|\theta_{u})$.  -->

<!-- Sachant que -->

<!-- \begin{align} -->

<!-- \label{loi_conditionel} -->

<!-- f_{\X|V}(x_i|v_i, \theta_y)&=|\frac{\partial \Y}{\partial \X}|f_\Y\bigg((\x_i-\gmu)\sqrt{v_i}\bigg)\\ -->

<!-- &=v_i^{\frac{p}{2}}\frac{1}{(2\pi)^{p/2}|\gSigma|^{1/2}}\exp\bigg(-\frac{v_i}{2}(\x_i-\gmu)^T\gSigma^{-1}(\x_i-\gmu)\bigg)    \\ -->

<!-- &=\frac{1}{(2\pi)^{p/2}|\frac{\gSigma}{v_i}|^{1/2}}\exp\bigg(-\frac{v_i}{2}(\x_i-\gmu)^T\gSigma^{-1}(\x_i-\gmu)\bigg)    \\ -->

<!-- \X|V &\sim \mathcal{N}^p(\gmu, \frac{1}{v_i}\gSigma) -->

<!-- \end{align} -->

<!-- Le passage au log aboutit à  -->

<!-- \begin{align} -->

<!-- \label{loi_conditionel} -->

<!-- \ln f_{\X|V}(x_i|v_i, \theta_y)&=\frac{p}{2}\ln v_i-\frac{p}{2}\ln 2\pi -\frac{1}{2} \ln |\gSigma| -\frac{v_i}{2}(\x_i-\gmu)^T\gSigma^{-1}(\x_i-\gmu)     -->

<!-- \end{align} -->

<!-- Les paramètres $\theta_y$ et $\theta_{v}$ peuvent être déterminés comme suit: -->

<!-- \begin{align} -->

<!-- \label{eq:lamda_x} -->

<!-- &\hat{\theta}_y=\arg \max_{\theta_y}\sum_{i=1}^N \mathbb{E}_{V|\X}\{\ln f_{\X|V}(\X_i|V_i,\theta_y)\big|\x_i,\theta'\} \\ -->

<!-- \label{eq:lamda_w} -->

<!-- &\hat{\theta}_v=\arg \max_{\theta_y}\sum_{i=1}^N \mathbb{E}_{V|\X}\{\ln f_{V}(V_i|\theta_{v})\big|\x_i,\theta'\}. -->

<!-- \end{align} -->

<!-- Substituer (\ref{loi_conditionel}) dans (\ref{eq:lamda_x}), différencier par rapport à $\gmu$ et $\gSigma$ et mettre le résultat à zéro donne: -->

<!-- \begin{align} -->

<!-- \sum_{i=1}^N \mathbb{E}_{V|\X}\{\frac{\partial}{\partial \gmu}\ln f_{\X|V}(\X_i|V_i,\theta_y)\big|\x_i,\theta'\}=0 =\sum_{i=1}^N \mathbb{E}_{V|\X}\{-\frac{V_i}{2}\frac{\partial}{\partial \gmu}(\X_i-\gmu)^T\gSigma^{-1}(\X_i-\gmu)\big|\x_i,\theta'\}\\ -->

<!-- \sum_{i=1}^N \mathbb{E}_{V|\X}\{\frac{\partial}{\partial \gSigma}\ln f_{\X|V}(\X_i|V_i,\theta_y)\big|\x_i,\theta'\}=0=\sum_{i=1}^N \mathbb{E}_{V|\X}\{-\frac{1}{2}\frac{\partial}{\partial \gSigma} \ln |\gSigma| -\frac{V_i}{2}\frac{\partial}{\partial \gSigma}(\X_i-\gmu)^T\gSigma^{-1}(\X_i-\gmu)    \big|\x_i,\theta'\}. -->

<!-- \end{align} -->

<!-- Le développement de ces équations aboutit  -->

<!-- \begin{align} -->

<!-- \sum_{i=1}^N \mathbb{E}_{V|\X}\{-\frac{V_i}{2}\frac{\partial}{\partial \gmu}\bigg[-\gmu^T\gSigma^{-1}\X_i -\X_i^T\gSigma^{-1}\gmu +\gmu^T\gSigma^{-1}\gmu\bigg]|\x_i,\theta'\}=0\\ -->

<!-- \sum_{i=1}^N \mathbb{E}_{V|\X}\{-\frac{1}{2}\frac{\partial}{\partial \gSigma} \ln |\gSigma| -\frac{V_i}{2}\frac{\partial}{\partial \gSigma}(\X_i-\gmu)^T\gSigma^{-1}(\X_i-\gmu)    \big|\x_i,\theta'\}=0. -->

<!-- \end{align} -->

<!-- sachant que  -->

<!-- \begin{align} -->

<!-- \frac{\partial}{\partial \gmu}[-\gmu^T\gSigma^{-1}\x_i] = -\gSigma^{-1}\x_i, \quad \frac{\partial}{\partial \gmu}[-\x_i^T\gSigma^{-1}\gmu] = -\gSigma^{-1}\x_i,~ \frac{\partial}{\partial \gmu}[\gmu^T\gSigma^{-1}\gmu]=2\gSigma^{-1}\gmu \text{~et~} \frac{\partial}{\partial \gSigma} \ln |\gSigma| = \gSigma^{-1}, \\ -->

<!-- \frac{\partial}{\partial \gSigma}(\x_i-\gmu)^T\gSigma^{-1}(\x_i-\gmu)= \frac{\partial}{\partial \gSigma}\mathrm{tr}\left(\gSigma^{-1}(\x_i-\gmu)(\x_i-\gmu)^T\right)= - \gSigma^{-1}(\x_i-\gmu)(\x_i-\gmu)^T\gSigma^{-1} -->

<!-- \end{align} -->

<!--We can show that the parameters $\boldsymbol{\mu}$ and $\Sigma$ are given by -->

$$
\hat{\boldsymbol{\mu}} = \frac{\sum_{i=1}^{N}\mathbf{x}_i\mathbb{E}_{V|\mathbf{X}}\{V_i|\mathbf{x}_i,\theta'\}}{\sum_{i=1}^{N}\mathbb{E}_{V|\mathbf{X}}\{V_i|\mathbf{x}_i,\theta'\}}
$$ {#eq-EM_K1}

$$
\hat{\boldsymbol{\Sigma}} = \frac{1}{N}\sum_{i=1}^{N}\mathbb{E}_{V|\mathbf{X}}\{V_i|\mathbf{x}_i,\theta'\}(\mathbf{x}_i-\boldsymbol{\mu})(\mathbf{x}_i-\boldsymbol{\mu})^T
$$ {#eq-EM_K2}

where $\mathbb{E}_{V|\mathbf{X}}\{V_i|\mathbf{x}_i,\theta'\}$ is the posterior expectation of V which depends on the values $\mathbf{x}_i$. The analytical expression is provided in the next subsection.

#### Estimation of $\theta_v=(\nu)$ {.unnumbered .unlisted}

The probability density function of $V$ is given by 
$$
f_{V}(v_i|\theta_{v}) = \big(\frac{\nu}{2}\big)^{\frac{\nu}{2}}\frac{1}{\Gamma(\nu/2)}v_i^{\frac{\nu}{2}-1}\exp{-\frac{\nu}{2}v_i}
$$ {#eq-v_distribution}

Substituting @eq-v_distribution into @eq-lamda_x, differentiating with respect to $ν$, and setting the result to zero provides the following result:
<!-- %************************************************************************* -->

<!-- \subsection{Estimation de $\theta_v=(\nu)$} -->

<!-- %************************************************************************* -->

<!-- la distribution de $V$ est alors  -->

<!-- \begin{align} -->

<!-- \label{loi_conditionel_2} -->

<!-- f_{V}(v_i|\theta_{v})&= (\frac{\nu}{2})^{\frac{\nu}{2}}\frac{1}{\Gamma(\nu/2)}v_i^{\frac{\nu}{2}-1}e^{-\frac{\nu}{2}v_i} -->

<!-- \end{align} -->

<!-- Il est clair que la distribution de $V=\frac{u}{\nu}\sim \text{Gamma}(\frac{\nu}{2}, \frac{\nu}{2})$. \\ -->

<!-- Le passage au log  -->

<!-- \begin{align} -->

<!-- \label{loi_conditionel_2} -->

<!-- \ln f_{V}(v_i|\theta_{v})&= \frac{\nu}{2} \ln\frac{\nu}{2} -\ln\Gamma(\nu/2)+(\frac{\nu}{2}-1)\ln v_i -\frac{v_i\nu}{2} -->

<!-- \end{align} -->

<!-- En substituant (\ref{loi_conditionel_2}) dans (\ref{eq:lamda_w}), en différenciant par rapport à $\nu$ et en mettant le résultat à zéro donne : -->

<!-- \begin{align} -->

<!-- \sum_{i=1}^N \mathbb{E}_{V|\X}\{\frac{\partial}{\partial \nu}\ln f_{V}(V_i|\theta_{v})|\x_i,\theta'\}&=0 = -->

<!-- \sum_{i=1}^N \mathbb{E}_{V|\X}\{1+\ln \frac{\nu}{2}-\psi(\frac{\nu}{2})+\ln V_i-V_i |\x_i,\theta'\}\\ -->

<!-- &=\sum_{i=1}^N 1+\ln \frac{\nu}{2}-\psi(\frac{\nu}{2})+ \mathbb{E}_{V|\X}\{\ln V_i|\x_i,\theta'\} -\mathbb{E}_{V|\X}\{V_i |\x_i,\theta'\} -->

<!-- \end{align} -->

<!-- En conclusion -->

$$
1+\ln \frac{\hat{\nu}}{2}-\psi(\frac{\hat{\nu}}{2})=\frac{1}{N}\sum_{i=1}^N \mathbb{E}_{V|\mathbf{X}}\{V_i |\mathbf{x}_i,\theta'\} -\mathbb{E}_{V|\mathbf{X}}\{\ln V_i|\mathbf{x}_i,\theta'\}
$$ {#eq-EM_G_1}

#### Expression of $\mathbb{E}_{V|\mathbf{X}}\{V_i |\mathbf{x}_i,\theta'\}$ and $\mathbb{E}_{V|\mathbf{X}}\{\ln V_i|\mathbf{x}_i,\theta'\}$ {.unnumbered .unlisted}

The posterior expectation $\mathbb{E}_{V|\mathbf{X}}\{V_i |\mathbf{x}_i,\theta'\}$ and $\mathbb{E}_{V|\mathbf{X}}\{\ln V_i|\mathbf{x}_i,\theta'\}$ depend on the posterior distribution $f_{V|\mathbf{X}}(v_i|\mathbf{x}_i)$ expressed as follows  

$$\begin{array}{lll}
f_{V|\mathbf{X}}(v_i|\mathbf{x}_i) & = &\displaystyle{\frac{f_{\mathbf{X}|V}(\mathbf{x_i}|v_i, \theta_y)f_{V}(v_i|\theta_{v})}{f_{\mathbf{X}}(\mathbf{x_i}|\theta_y, \theta_v)}}\\
&= & \displaystyle{\frac{v_i^{\frac{p+\nu}{2}-1}}{\Gamma(\frac{p+\nu}{2})} (\frac{1}{2})^{\frac{p+\nu}{2}}}\bigg[\nu+(\mathbf{x}_i-\boldsymbol{\mu})^T\boldsymbol{\Sigma}^{-1}(\mathbf{x}_i-\boldsymbol{\mu})\bigg]^{\frac{\nu+p}{2}} \exp-\frac{v_i}{2}\bigg[\nu+(\mathbf{x}_i-\boldsymbol{\mu})^T\boldsymbol{\Sigma}^{-1}(\mathbf{x}_i-\boldsymbol{\mu})\bigg]
\end{array}
$$
It is worth noticing that $V|\mathbf{X}\sim Gamma\left(\frac{p+\nu}{2},\nu+(\mathbf{x}_i-\boldsymbol{\mu})^T\boldsymbol{\Sigma}^{-1}(\mathbf{x}_i-\boldsymbol{\mu})\right)$.
Consequently, the posterior expectations are given by:

$$
\mathbb{E}_{V|\mathbf{X}}\{V_i|\mathbf{x}_i,\theta'\} = \frac{\nu+p}{\nu+(\mathbf{x}_i-\boldsymbol{\mu})^T\boldsymbol{\Sigma}^{-1}(\mathbf{x}_i-\boldsymbol{\mu})}
$$ {#eq-E1}

and:

$$
\mathbb{E}_{V|\mathbf{X}}\{\ln V_i|\mathbf{x}_i,\theta'\} = \psi(\frac{\nu+p}{2}) -\ln \frac{1}{2}\bigg[\nu+(\mathbf{x}_i-\boldsymbol{\mu})^T\boldsymbol{\Sigma}^{-1}(\mathbf{x}_i-\boldsymbol{\mu})\bigg]
$$ {#eq-E2}

<!-- \begin{tcolorbox}[arc=1ex, colback=yellow, colframe=yellow, left=3pt, right=3pt, top=2pt, bottom=2pt] -->

<!-- \begin{align} -->

<!-- \mathbb{E}_{V|\X}\{V_i |\x_i,\theta'\}&=\frac{\nu+p}{\nu+(\x_i-\gmu)^T\gSigma^{-1}(\x_i-\gmu)} -->

<!-- \end{align} -->

<!-- \end{tcolorbox} -->

<!-- %************************************************************************************************* -->

<!-- \subsection{Détermination de $\mathbb{E}_{V|\X}\{V_i |\x_i,\theta'\}$ et $\mathbb{E}_{V|\X}\{\ln V_i|\x_i,\theta'\}$} -->

<!-- %************************************************************************************************* -->

<!-- L'espérance a posteriori $\mathbb{E}_{V|\X}\{V_i |\x_i,\theta'\}$ et $\mathbb{E}_{V|\X}\{\ln V_i|\x_i,\theta'\}$ dépendent de la densité a posteriori $f_{V|\X}(v_i|\x_i)$ qui est exprimé comme suit: -->

<!-- \begin{align} -->

<!-- f_{V|\X}(v_i|\x_i)&=\frac{f_{\X|V}(\x_i|v_i,\theta_{y})f_V(v_i|\theta_{v})}{f_{\X}(\x_i|\theta_{y},\theta_{v})}\\ -->

<!-- &=\frac{v_i^{\frac{p}{2}}\frac{1}{(2\pi)^{p/2}|\gSigma|^{1/2}}\exp\bigg[-\frac{v_i}{2}(\x_i-\gmu)^T\gSigma^{-1}(\x_i-\gmu)\bigg] (\frac{\nu}{2})^{\frac{\nu}{2}}\frac{1}{\Gamma(\nu/2)}v_i^{\frac{\nu}{2}-1}e^{-\frac{\nu}{2}v_i}       }{ \frac{\Gamma(\frac{\nu+p}{2})}{\Gamma(\nu/2)} \frac{1}{\pi^{p/2}|\gSigma|^{1/2}\nu^{\frac{p}{2}}} \bigg[1+\frac{1}{\nu}(\x_i-\gmu)^T\gSigma^{-1}(\x_i-\gmu)\bigg]^{-\frac{\nu+p}{2}}   }   \\ -->

<!-- &=\frac{v_i^{\frac{p+\nu}{2}-1}}{\Gamma(\frac{p+\nu}{2})}(\frac{\nu}{2})^{\frac{p+\nu}{2}}\bigg[1+\frac{1}{\nu}(\x_i-\gmu)^T\gSigma^{-1}(\x_i-\gmu)\bigg]^{\frac{\nu+p}{2}} \exp-\frac{v_i}{2}\bigg[\nu+(\x_i-\gmu)^T\gSigma^{-1}(\x_i-\gmu)\bigg] -->

<!-- \end{align} -->

<!-- \begin{tcolorbox}[arc=1ex, colback=yellow, colframe=yellow, left=3pt, right=3pt, top=2pt, bottom=2pt] -->

<!-- \begin{align} -->

<!-- V|\X &\sim \text{Gamma}\bigg(\frac{\nu+p}{2}, \frac{1}{2}\bigg[\nu+(\x_i-\gmu)^T\gSigma^{-1}(\x_i-\gmu)\bigg] \bigg) -->

<!-- \end{align} -->

<!-- \end{tcolorbox} -->

<!-- Nous pouvons maintenant exprimé l'espérance  -->

<!-- \begin{tcolorbox}[arc=1ex, colback=yellow, colframe=yellow, left=3pt, right=3pt, top=2pt, bottom=2pt] -->

<!-- \begin{align} -->

<!-- \mathbb{E}_{V|\X}\{V_i |\x_i,\theta'\}&=\frac{\nu+p}{\nu+(\x_i-\gmu)^T\gSigma^{-1}(\x_i-\gmu)} -->

<!-- \end{align} -->

<!-- \end{tcolorbox} -->

<!-- L'expression de $\mathbb{E}_{V|\X}\{\ln V_i|\x_i,\theta'\}$ est donnée  -->

<!-- \begin{tcolorbox}[arc=1ex, colback=yellow, colframe=yellow, left=3pt, right=3pt, top=2pt, bottom=2pt] -->

<!-- \begin{align} -->

<!-- \mathbb{E}_{V|\X}\{\ln V_i|\x_i,\theta'\}&= \psi(\frac{\nu+p}{2}) -\ln \frac{1}{2}\bigg[\nu+(\x_i-\gmu)^T\gSigma^{-1}(\x_i-\gmu)\bigg]  -->

<!-- \end{align} -->

<!-- \end{tcolorbox} -->

<!-- \begin{table}[h] -->

<!-- \begin{tabular}{l} -->

<!-- \toprule -->

#### Algorithm {#sec-mtd_algo .unnumbered .unlisted}

From these expressions, we can build an algorithm for the parameter
estimation:

<!-- \textbf{Algorithme }  \\ -->

<!-- \bottomrule -->

<!-- \end{tabular} -->

<!-- \end{table} -->

<!-- **Input**: $N, \mathbf{x}_i, \epsilon$ -->

<!-- **Output**: $\hat{\theta}=(\hat{\boldsymbol{\mu}},\hat{\Sigma},\hat{\nu})$ -->

<!-- *   [**Initialisation:**]{.underline} -->

1.  Initialization of the parameters
    $\theta'=({\boldsymbol{\mu}}',{\boldsymbol{\Sigma}}',{\nu}')$

2.  Repeat

    -   Compute $\mathbb{E}_{V|\mathbf{X}}\{V_i |\mathbf{x}_i,\theta\}$ using 
        @eq-E1 and $\mathbb{E}_{V|\mathbf{X}}\{\ln V_i|\mathbf{x}_i,\theta\}$
        using @eq-E2

    -   Estimate $\hat{\boldsymbol{\mu}}$ using @eq-EM_K1

    -   Estimate $\hat{\boldsymbol{\Sigma}}$ using @eq-EM_K2

    -   Resolve @eq-EM_G_1 to estimate $\hat{\nu}$
    
    -   Compute the stop criterion: $D\leftarrow\Vert\hat{\theta}-\theta'\Vert$
    
    -   Define the inputs for the next iteration: $\boldsymbol{\mu}',\boldsymbol{\Sigma}',\nu' \leftarrow\hat{\boldsymbol{\mu}},\hat{\boldsymbol{\Sigma}},\hat{\nu}$
    
    -   **Until** $D<\epsilon$

<!--           *   Compute the stop criterion: $D\leftarrow\Vert\hat{\theta}-\theta'\Vert$ -->

<!--           *   Define the inputs for the next iteration: $\boldsymbol{\mu}',\Sigma',\nu' \leftarrow\hat{\boldsymbol{\mu}},\hat{\Sigma},\hat{\nu}$ -->

<!--     1.    **Until** $D<\epsilon$ -->

3.   **Return** $(\hat{\boldsymbol{\mu}},\hat{\boldsymbol{\Sigma}},\hat{\nu})$ 

### Rényi divergence

Let $\mathbf{X}^1$ and $\mathbf{X}^2$ be two random vectors of
$\mathbb{R}^p$ from MTDs with probability density functions given as follows: $$
    \displaystyle{ f_{\mathbf{X}^1}(\mathbf{x}|\nu_1, \boldsymbol{\Sigma}_1) = f_{\mathbf{X}^1}(\mathbf{x}|\nu_1, \boldsymbol{0}, \boldsymbol{\Sigma}_1) = \frac{\Gamma\left( \frac{\nu_1+p}{2} \right) |\boldsymbol{\Sigma}_1|^{-1/2}}{\Gamma\left( \frac{\nu_1}{2} \right) (\nu_1 \pi)^{p/2}} \left( 1 + \frac{1}{\nu_1} \mathbf{x}^T \boldsymbol{\Sigma}_1^{-1} \mathbf{x} \right)^{-\frac{\nu_1+p}{2}} }
$$ and $$
    \displaystyle{ f_{\mathbf{X}^2}(\mathbf{x}|\nu_2, \boldsymbol{\Sigma}_2) = f_{\mathbf{X}^2}(\mathbf{x}|\nu_2, \boldsymbol{0}, \boldsymbol{\Sigma}_2) = \frac{\Gamma\left( \frac{\nu_2+p}{2} \right) |\boldsymbol{\Sigma}_2|^{-1/2}}{\Gamma\left( \frac{\nu_2}{2} \right) (\nu_2 \pi)^{p/2}} \left( 1 + \frac{2}{\nu_2} \mathbf{x}^T \boldsymbol{\Sigma}_2^{-1} \mathbf{x} \right)^{-\frac{\nu_2+p}{2}} }
$$

The Rényi divergence (RD) of order $\beta$ ($\beta >0$, $\beta \neq 1$) between
$\mathbf{X}^1$ and $\mathbf{X}^2$ is: $$\begin{array}{lll}
\displaystyle{D_R^{\beta}(\mathbf{X}^1 \Vert \mathbf{X}^2)} &   = &   \displaystyle{\frac{1}{\beta-1} \ln\left({\mathbb{E}_{X^1}\left[ \left( \frac{f_{\mathbf{X}^1}(\mathbf{x}|\nu_1, \boldsymbol{\Sigma}_1)}{f_{\mathbf{X}^2}(\mathbf{x}|\nu_2, \boldsymbol{\Sigma}_2)} \right)^{\beta-1} \right]}\right)} \\
\\
& =   &    \displaystyle{\frac{1}{\beta-1} \ln\left( \int_{\mathbb{R}^p}{\left( \frac{f_{\mathbf{X}^1}(\mathbf{x}|\nu_1, \boldsymbol{\Sigma}_1)}{f_{\mathbf{X}^2}(\mathbf{x}|\nu_2, \boldsymbol{\Sigma}_2)} \right)^{\beta-1} f_{\mathbf{X}^1}(\mathbf{x}|\nu_1, \boldsymbol{\Sigma}_1) \mathrm{d}\mathbf{x}} \right)}
\end{array}$$ {#eq-rd}

Let $\lambda_1, \dots, \lambda_p$ be the eigenvalues of the square
matrix $\boldsymbol{\Sigma}_1 \boldsymbol{\Sigma}_2^{-1}$ sorted in increasing order:
$\lambda_1 < \dots < \lambda_p$. Let $\delta_1 = \frac{\nu_1+p}{2}\beta$
and $\delta_2 = \frac{\nu_2+p}{2}(1-\beta)$.

@bouhlel23 shows that the RD of order $\beta$ between $\mathbf{X}^1$ and
$\mathbf{X}^2$ is given by:

$$\begin{array}{lll}
\displaystyle{D_R^{\beta}(\mathbf{X}^1 \Vert \mathbf{X}^2)} &   = &   \displaystyle{\frac{1}{1-\beta} \left[ \beta \ln\left( \frac{\Gamma\left(\frac{\nu_1+p}{2}\right) \Gamma\left(\frac{\nu_2}{2}\right) \nu_2^\frac{p}{2}}{\Gamma\left(\frac{\nu_2+p}{2}\right) \Gamma\left(\frac{\nu_1}{2}\right) \nu_1^\frac{p}{2}} \right) + \ln\left( \frac{\Gamma\left(\frac{\nu_2+p}{2}\right)}{\Gamma\left(\frac{\nu_2}{2}\right)} \right) \right.} \\
\\
&&   \displaystyle{\left. + \ln\left( \frac{\Gamma\left(\delta_1 + \delta_2 - \frac{p}{2}\right)}{\Gamma(\delta_1 + \delta_2)} \right) - \frac{\beta}{2} \sum_{i=1}^p{\ln\lambda_i} + \ln F_D \right]}
\end{array}$$ {#eq-rdmtd}

where $F_D$ depends on the values of
$\frac{\nu_1}{\nu_2}\lambda_1, \dots, \frac{\nu_1}{\nu_2}\lambda_p$ :

-   If $\frac{\nu_1}{\nu_2}\lambda_1 > 1$,

$$
\displaystyle{ F_D = F_D^{(p)}\bigg( \delta_1, \underbrace{\frac{1}{2}, \dots, \frac{1}{2}}_p; \delta_1 + \delta_2; 1 - \frac{\nu_2}{\nu_1}\frac{1}{\lambda_1}, \dots, 1 - \frac{\nu_2}{\nu_1}\frac{1}{\lambda_p} \bigg) }
$$

-   If $\frac{\nu_1}{\nu_2}\lambda_p < 1$,

$$
\displaystyle{ F_D = \prod_{i=1}^p{\left(\frac{\nu_1}{\nu_2}\lambda_i\right)^\frac{1}{2}} F_D^{(p)}\bigg( \delta_2, \underbrace{\frac{1}{2}, \dots, \frac{1}{2}}_p; \delta_1 + \delta_2; 1 - \frac{\nu_1}{\nu_2}\lambda_1, \dots, 1 - \frac{\nu_1}{\nu_2}\lambda_p \bigg) }
$$

-   If $\frac{\nu_1}{\nu_2}\lambda_1 < 1 < \frac{\nu1}{\nu_2}\lambda_p$,

$$\begin{array}{lll}
F_D &   = &   \displaystyle{ \left(\frac{\nu_2}{\nu_1}\frac{1}{\lambda_p}\right)^{\delta_2} \prod_{i=1}^p{\left(\frac{\nu_1}{\nu_2}\lambda_i\right)^\frac{1}{2}} } \\
&&    \displaystyle{ \times F_D^{(p)}\bigg( \delta_2, \underbrace{\frac{1}{2}, \dots, \frac{1}{2}, \delta_1 + \delta_2 - \frac{p}{2}}_p; \delta_1 + \delta_2; 1 - \frac{\lambda_1}{\lambda_p}, \dots, 1 - \frac{\lambda_{p-1}}{\lambda_p}, 1 - \frac{\nu_2}{\nu_1}\frac{1}{\lambda_p} \bigg) }
\end{array}
$$

where $F_D^{(n)}$ is the Lauricella $D$-hypergeometric function defined
by: $$
    F_D^{(n)}\left(a; b_1, \dots, b_n; c; x_1, \dots, x_n\right) = \sum_{m_1=0}^\infty \dots \sum_{m_n=0}^\infty{\frac{(a)_{m_1+\dots+m_n} (b_1)_{m_1} \dots (b_n)_{m_n}}{(c)_{m_1+\dots+m_n}} \ \frac{{x_1}^{m_1}}{m_1!} \dots \frac{{x_n}^{m_n}}{m_n!}}
$$ {#eq-lauric} and the Pochhammer symbol $(q)_i$, where $q$ is a real
and $i$ is a non-negative integer, is given by: $$
\displaystyle{ (q)_i = \frac{\Gamma(q+i)}{\Gamma(q)} =} \left\{ \begin{array}{ll}
1 & \text{if } i = 0 \\
\displaystyle{ q (q+1) \dots (q+i-1) } & \text{if } i \geq 1
\end{array} \right.
$$ {#eq-pochhammer}

### Bhattacharyya and Hellinger divergences

The Bhattacharyya divergence is computed from the RD of order
$\beta = \frac{1}{2}$: $$
D_B(\mathbf{X}^1 \Vert \mathbf{X}^2) = \frac{1}{2} D_R^\frac{1}{2}(\mathbf{X}^1 \Vert \mathbf{X}^2)
$$ {#eq-bdmtd}

And the Hellinger divergence: $$
D_H(\mathbf{X}^1 \Vert \mathbf{X}^2) = 1 - \exp\left(-\frac{1}{2} D_R^\frac{1}{2}(\mathbf{X}^1 \Vert \mathbf{X}^2)\right)
$$ {#eq-hdmtd}

### Kullback-Leibler divergence {#sec-kldt}

With the same notations, the KLD of $\mathbf{X}_1$ from $\mathbf{X}_2$
is given by:

$$\begin{array}{lll}
D_{KL}(\mathbf{X}^1 \Vert \mathbf{X}^2) &   = &   \displaystyle{ \ln\left( \frac{\Gamma\left(\frac{\nu_1+p}{2}\right) \Gamma\left(\frac{\nu_2}{2}\right) \nu_2^\frac{p}{2}}{\Gamma\left(\frac{\nu_2+p}{2}\right) \Gamma\left(\frac{\nu_1}{2}\right) \nu_1^\frac{p}{2}} \right) + \frac{\nu_2 - \nu_1}{2} \left( \psi\left(\frac{\nu_1+p}{2}\right) - \psi\left(\frac{\nu_1}{2}\right) \right) } \\
&&    \displaystyle{ - \frac{1}{2}\sum_{i=p}^p{\ln\lambda_i} - \frac{\nu_2+p}{2} \times D }
\end{array}
$$ {#eq-klmtd}

where $D$ depends on the values of
$\frac{\nu_1}{\nu_2}\lambda_1, \dots, \frac{\nu_1}{\nu_2}\lambda_p$:

-   If $\displaystyle{\frac{\nu_1}{\nu_2}\lambda_1 > 1}$,
    $$\begin{array}{lll}
    D &   = &   \displaystyle{ \prod_{i=1}^p{\left(\frac{\nu_2}{\nu_1}\frac{1}{\lambda_i}\right)^\frac{1}{2}} \frac{\partial}{\partial{a}}{ \bigg\{ F_D^{(p)}\bigg( \frac{\nu_1+p}{2}, \underbrace{\frac{1}{2}, \dots, \frac{1}{2}}_p; a + \frac{\nu_1+p}{2}; 1 - \frac{\nu_2}{\nu_1}\frac{1}{\lambda_1}, \dots, 1 - \frac{\nu_2}{\nu_1}\frac{1}{\lambda_p} \bigg) \bigg\} } \bigg|_{a=0} } \\
    &   = &   \displaystyle{ \prod\limits_{i=1}^p\left(\frac{\nu_2}{\nu_1}\frac{1}{\lambda_i}\right)^{\frac{1}{2}}\sum\limits_{m_1\geq 0}...\sum\limits_{m_p\geq 0} \bigg(\frac{\nu_1+p}{2}\bigg)_{\sum m_i}\frac{\partial}{\partial a}\bigg\lbrace \frac{1}{\big(a+\frac{\nu_1+p}{2}\big)_{\sum m_i}}  \bigg\rbrace\bigg|_{a=0} } \\
    &&    \displaystyle{ \times \bigg(\frac{1}{2}\bigg)_{m_1}...\bigg(\frac{1}{2}\bigg)_{m_p} \frac{1}{m_1!}\bigg(1-\frac{\nu_2}{\nu_1}\frac{1}{\lambda_1}\bigg)^{m_1} \dots \frac{1}{m_p!}\bigg(1-\frac{\nu_2}{\nu_1}\frac{1}{\lambda_p}\bigg)^{m_p} }
    \end{array}$$ {#eq-klmtd1} where: $$
    \displaystyle{\frac{\partial}{\partial a}\bigg\lbrace \frac{1}{\left(a+\frac{\nu_1+p}{2}\right)_{\sum_{i=1}^p m_i}}  \bigg\rbrace\bigg|_{a=0} = \frac{-1}{\big(\frac{\nu_1+p}{2}\big)_{\sum_{i=1}^pm_i}} \sum_{k=0}^{\sum m_i -1} \frac{1}{k+\frac{\nu_1+p}{2}}}
    $$

-   If $\displaystyle{\frac{\nu_1}{\nu_2}\lambda_p < 1}$,
    $$\begin{array}{lll}
    D &   = &   \displaystyle{ \frac{\partial}{\partial{a}}{ \bigg\{ F_D^{(p)}\bigg( a, \underbrace{\frac{1}{2}, \dots, \frac{1}{2}}_p; a + \frac{\nu_1+p}{2}; 1 - \frac{\nu_1}{\nu_2}\lambda_1, \dots, 1 - \frac{\nu_1}{\nu_2}\lambda_p \bigg) \bigg\} } \bigg|_{a=0} } \\
    &   = &   \displaystyle{ \sum_{m_1\geq 0} \dots \sum_{m_p\geq 0} \frac{\partial}{\partial a}\bigg\lbrace \frac{(a)_{\sum mi}}{\big(a+\frac{\nu_1+p}{2}\big)_{\sum m_i}}  \bigg\rbrace\bigg|_{a=0}  \bigg(\frac{1}{2}\bigg)_{m_1}...\bigg(\frac{1}{2}\bigg)_{m_p} } \\
    &&    \displaystyle{ \times \frac{1}{m_1!}\bigg(1-\frac{\nu_1}{\nu_2}\lambda_1\bigg)^{m_1} \dots \frac{1}{m_p!}\bigg(1-\frac{\nu_1}{\nu_2}\lambda_p\bigg)^{m_p} }
    \end{array}$$ {#eq-klmt2} where: $$
    \displaystyle{ \frac{\partial}{\partial a}\bigg\lbrace \frac{(a)_{\sum m_i}}{\big(a+\frac{\nu_1+p}{2}\big)_{\sum_{i=1}^p m_i}}  \bigg\rbrace\bigg|_{a=0} = \frac{(1)_{\sum m_i} }{\big(\frac{\nu_1+p}{2}\big)_{\sum m_i}}\frac{1}{\sum m_i} }
    $$

-   If
    $\displaystyle{\frac{\nu_1}{\nu_2}\lambda_1 < 1 < \frac{\nu1}{\nu_2}\lambda_p}$,
    $$\begin{array}{lll}
    D &   = &   \displaystyle{ -\ln\left(\frac{\nu_1}{\nu_2}\lambda_p\right) } + \\
    &&    \displaystyle{ \frac{\partial}{\partial{a}}{ \bigg\{ F_D^{(p)}\bigg( a, \underbrace{\frac{1}{2}, \dots, \frac{1}{2}, a + \frac{\nu_1}{2}}_p; a + \frac{\nu_1+p}{2}; 1 - \frac{\lambda_1}{\lambda_p}, \dots, 1 - \frac{\lambda_{p-1}}{\lambda_p}, 1 - \frac{\nu_2}{\nu_1}\frac{1}{\lambda_p} \bigg) \bigg\} } \bigg|_{a=0} } \\
    &   = &   \displaystyle{ -\ln\bigg(\frac{\nu_1}{\nu_2}\lambda_p\bigg) +\sum\limits_{m_1\geq 0} \dots \sum_{m_p\geq 0} \frac{\partial}{\partial a}\bigg\lbrace \frac{(a)_{\sum mi} (a+\frac{\nu_1}{2})_{m_p}}{\big(a+\frac{\nu_1+p}{2}\big)_{\sum m_i}}  \bigg\rbrace\bigg|_{a=0} } \\
    &&    \displaystyle{ \times \bigg(\frac{1}{2}\bigg)_{m_1} \dots \bigg(\frac{1}{2}\bigg)_{m_{p-1}} \frac{1}{m_1!}\left(1-\frac{\lambda_1}{\lambda_p}\right)^{m_1} \dots  \frac{1}{m_p!}\left(1-\frac{\nu_2}{\nu_1}\frac{1}{\lambda_p}\right)^{m_p} }
    \end{array}$$ {#eq-klmtd3} where: $$
    \displaystyle{ \frac{\partial}{\partial a}\bigg\lbrace \frac{(a)_{\sum m_i}(a+\frac{\nu_1}{2})_{m_p}}{\big(a+\frac{\nu_1+p}{2}\big)_{\sum_{i=1}^p m_i}}  \bigg\rbrace\bigg|_{a=0} = \frac{(1)_{\sum m_i} }{\big(\frac{\nu_1+p}{2}\big)_{\sum m_i}}\frac{(\frac{\nu_1}{2})_{m_p}}{\sum m_i} }
    $$

### Study of the divergence

The divergence between two multivariate $t$ distributions
$\mathbf{X}^1 \sim f_{\mathbf{X}^1}(\mathbf{x}|\nu_1, \boldsymbol{0}, \boldsymbol{\Sigma}_1)$
and
$\mathbf{X}^2 \sim f_{\mathbf{X}^2}(\mathbf{x}|\nu_2, \boldsymbol{0}, \boldsymbol{\Sigma}_2)$
depends on the values of the parameters of these distributions.

In the examples below, the divergences between $\mathbf{X}^1$ and $\mathbf{X}^2$
are computed using the functions `dlstudent` and `kld` from the
[`multvardiv`](https://CRAN.R-project.org/package=multvardiv) package.

#### Among the degrees of freedom

Let $\mathbf{X}^1 \sim MTD(\mathbf{0}, \nu_1, \boldsymbol{\Sigma})$
and $\mathbf{X}^2 \sim MTD(\mathbf{0}, \nu_2, \boldsymbol{\Sigma})$

where $\nu_1$ take values in $]0 ; 30]$, $\nu_2 = 2$, and
$\boldsymbol{\Sigma} = \left(\begin{array}{cc}
1 & 0.3 \\
0.3 & 1
\end{array}\right)$.

The results are showed in @fig-divnu1.
Obviously, for each of these divergences,
$D(X^1 \parallel X^2) = 0$ when $\nu_1 = \nu_2$.
For $\nu_1 \in ]0; \nu_2]$, the divergence decreases quickly to 0,
then for $\nu_1 \in ]\nu_2; +\infty]$, it increases slowly and converges
to a limit.

```{r}
#| cache: true
#| warning: false

nu <- 2#5

# Values of nu1
nu1 <- c(seq(0.1, nu, by = 0.1), seq(nu+1, 50, by = 2))

# Fixed values of the other parameters
nu2 <- nu
Sigma <- matrix(c(1, 0.3, 0.3, 1), nrow = 2)

# Values of lambda*nu1/nu2
nulambda <- nu1/nu2

# Rényi divergence, beta = 0.25
b <- 0.25
renyi_0_25 <- sapply(nu1, function(nu) {
  diststudent(nu1 = nu, Sigma1 = Sigma,
              nu2 = nu2, Sigma2 = Sigma,
              dist = "renyi", bet = b)
})

# Rényi divergence, beta = 1.5
b <- 1.5
renyi_1_5 <- sapply(nu1, function(nu) {
  diststudent(nu1 = nu, Sigma1 = Sigma,
              nu2 = nu2, Sigma2 = Sigma,
              dist = "renyi", bet = b)
})

# Battacharyya divergence
bhattacharyya <- sapply(nu1, function(nu) {
  diststudent(nu1 = nu, Sigma1 = Sigma,
              nu2 = nu2, Sigma2 = Sigma,
              dist = "bhattacharyya")
})

# Hellinger divergence
hellinger <- sapply(nu1, function(nu) {
  diststudent(nu1 = nu, Sigma1 = Sigma,
              nu2 = nu2, Sigma2 = Sigma,
              dist = "hellinger")
})

# Kullback-Leibler divergence
kullback_leibler <- sapply(nu1, function(nu) {
  kld(Sigma1 = Sigma, Sigma2 = Sigma,
      nu1 = nu, nu2 = nu2,
      distribution = "mtd")
})

# Table of the results
divnu <- data.frame(
  nu1 = nu1,
  nulambda = nulambda,
  renyi_0_25 = renyi_0_25,
  renyi_1_5 = renyi_1_5,
  bhattacharyya = bhattacharyya,
  hellinger = hellinger,
  kullback_leibler = kullback_leibler
)
```

<!-- $\mathbf{X}^1$ et $\mathbf{X}^2$ -->

```{r}
#| label: fig-divnu1
#| fig-cap: "Divergence between the distributions $\\mathbf{X}^1$ et $\\mathbf{X}^2$"
#| fig-subcap:
#|    - "Renyi ($\\beta = 0.25$)"
#|    - "Renyi ($\\beta = 1.5$)"
#|    - ""
#|    - "Bhattacharyya"
#|    - "Hellinger"
#|    - "Kullback-Leibler"
#| layout-ncol: 3
#| layout-nrow: 2
plot(renyi_0_25~nu1, data = divnu, type = "b", pch = 16,
     xlab = substitute(nu[1]), ylab = "Divergence")
axis(side = 1, at = nu2, labels = substitute(nu[2]), cex.axis = 1.5)

plot(renyi_1_5~nu1, data = divnu, type = "b", pch = 16,
     xlab = substitute(nu[1]), ylab = "Divergence")
axis(side = 1, at = nu2, labels = substitute(nu[2]), cex.axis = 1.5)

plot(0, 0, type = "n", bty = "n", xaxt = "n", yaxt = "n", xlab = "", ylab = "")

plot(bhattacharyya~nu1, data = divnu, type = "b", pch = 16,
     xlab = substitute(nu[1]), ylab = "Divergence")
axis(side = 1, at = nu2, labels = substitute(nu[2]), cex.axis = 1.5)

plot(hellinger~nu1, data = divnu, type = "b", pch = 16,
     xlab = substitute(nu[1]), ylab = "Divergence")
axis(side = 1, at = nu2, labels = substitute(nu[2]), cex.axis = 1.5)

plot(kullback_leibler~nu1, data = divnu, type = "b", pch = 16,
     xlab = substitute(nu[1]), ylab = "Divergence")
axis(side = 1, at = nu2, labels = substitute(nu[2]), cex.axis = 1.5)
```

#### Among the scale matrix: first situation

Let $\mathbf{X}^1 \sim MTD(\mathbf{0}, \nu, \boldsymbol{\Sigma}^1)$
and $\mathbf{X}^2 \sim MTD(\mathbf{0}, \nu, \boldsymbol{\Sigma}^2)$

where $\nu = 1$,
$\boldsymbol{\Sigma}^2 = \left(\begin{array}{cc}
3 & 0.3 \\
0.3 & 1
\end{array}\right)$  
and $\boldsymbol{\Sigma}^1 = \left(\begin{array}{cc}
s_1 & 0.3 \\
0.3 & 1
\end{array}\right)$, $s_1 \in ]0 ; 50]$.

The results are showed in @fig-divsigma11.
For each of these divergences,
$D(X^1 \parallel X^2) = 0$ when $s_1 = \boldsymbol{\Sigma}^2_{1, 1}$.
For $s_1 \in ]0; \boldsymbol{\Sigma}^2_{1, 1}]$, the divergence decreases quickly to 0,
then for $s_1 > \boldsymbol{\Sigma}^2_{1, 1}$, it increases slowly and converges.

```{r}
#| cache: true

s <- 3

# Values of Sigma1[1, 1]
s1 <- c(seq(0.1, s, by = 0.05), seq(s+1, 70, by = 1))

# The other parameters
nu <- 1
Sigma1 <- Sigma2 <- matrix(c(s, 0.3, 0.3, 1), nrow = 2)
Sigma1[1, 1] <- NA

# Values of lambda*nu1/nu2
invS2 <- solve(Sigma2)
lambdanu <- sapply(s1, function(s) {
  Sigma1[1, 1] <- s
  sort(eigen(Sigma1 %*% invS2, only.values = TRUE)$values, decreasing = FALSE)
})
ln1 <- lambdanu[1, ]
lnp <- lambdanu[2, ]

# Renyi divergence, beta = 0.25
b <- 0.25
renyi_0_25 <- sapply(s1, function(s1) {
  Sigma1[1, 1] <- s1
  diststudent(nu1 = nu, Sigma1 = Sigma1, nu2 = nu, Sigma2 = Sigma2,
              dist = "renyi", bet = b)
})

# Renyi divergence, beta = 1.5
b <- 1.5
renyi_1_5 <- sapply(s1, function(s1) {
  Sigma1[1, 1] <- s1
  diststudent(nu1 = nu, Sigma1 = Sigma1, nu2 = nu, Sigma2 = Sigma2,
              dist = "renyi", bet = b)
})

# Battacharyya divergence
bhattacharyya <- sapply(s1, function(s1) {
  Sigma1[1, 1] <- s1
  diststudent(nu1 = nu, Sigma1 = Sigma1, nu2 = nu, Sigma2 = Sigma2,
              dist = "bhattacharyya")
})

# Hellinger divergence
hellinger <- sapply(s1, function(s1) {
  Sigma1[1, 1] <- s1
  diststudent(nu1 = nu, Sigma1 = Sigma1, nu2 = nu, Sigma2 = Sigma2,
              dist = "hellinger")
})

# Kullback-Leibler divergence
kullback_leibler <- sapply(s1, function(s1) {
  Sigma1[1, 1] <- s1
  kld(Sigma1 = Sigma1, Sigma2 = Sigma2, nu1 = nu, nu2 = nu,
      distribution = "mtd")
})

# Table of the results
divS1 <- data.frame(
  s1 = s1, ln1 = ln1, lnp = lnp,
  renyi_0_25 = renyi_0_25,
  renyi_1_5 = renyi_1_5,
  bhattacharyya = bhattacharyya,
  hellinger = hellinger,
  kullback_leibler = kullback_leibler
)
```

<!-- $\mathbf{X}^1$ et $\mathbf{X}^2$ -->

```{r}
#| label: fig-divsigma11
#| fig-cap: "Divergence between the distributions $\\mathbf{X}^1$ et $\\mathbf{X}^2$"
#| fig-subcap:
#|    - "$\\lambda_1 \\frac{\\nu_1}{\\nu_2}$"
#|    - "$\\lambda_p \\frac{\\nu_1}{\\nu_2}$"
#|    - ""
#|    - "Renyi ($\\beta = 0.25$)"
#|    - "Renyi ($\\beta = 1.5$)"
#|    - ""
#|    - "Bhattacharyya"
#|    - "Hellinger"
#|    - "Kullback-Leibler"
#| layout-ncol: 3
#| layout-nrow: 2

plot(ln1~s1, data = divS1, type = "b", pch = 16,
     xlab = substitute(Sigma[1][1]), ylab = "lambda1 * nu1/nu2")
axis(side = 1, at = Sigma2[1, 1], labels = substitute(Sigma[1][1]), cex.axis = 1.5)

plot(lnp~s1, data = divS1, type = "b", pch = 16,
     xlab = substitute(Sigma[1][1]), ylab = "lambdap * nu1/nu2")
axis(side = 1, at = Sigma2[1, 1], labels = substitute(Sigma[1][1]), cex.axis = 1.5)

plot(0, 0, type = "n", bty = "n", xaxt = "n", yaxt = "n", xlab = "", ylab = "")
plot(renyi_0_25~s1, data = divS1, type = "b", pch = 16,
     xlab = substitute(Sigma[1][1]), ylab = "Divergence")
axis(side = 1, at = Sigma2[1, 1], labels = substitute(Sigma[1][1]), cex.axis = 1.5)

plot(renyi_1_5~s1, data = divS1, type = "b", pch = 16,
     xlab = substitute(Sigma[1][1]), ylab = "Divergence")
axis(side = 1, at = Sigma2[1, 1], labels = substitute(Sigma[1][1]), cex.axis = 1.5)

plot(0, 0, type = "n", bty = "n", xaxt = "n", yaxt = "n", xlab = "", ylab = "")
plot(bhattacharyya~s1, data = divS1, type = "b", pch = 16,
     xlab = substitute(Sigma[1][1]), ylab = "Divergence")
axis(side = 1, at = Sigma2[1, 1], labels = substitute(Sigma[1][1]), cex.axis = 1.5)

plot(hellinger~s1, data = divS1, type = "b", pch = 16,
     xlab = substitute(Sigma[1][1]), ylab = "Divergence")
axis(side = 1, at = Sigma2[1, 1], labels = substitute(Sigma[1][1]), cex.axis = 1.5)

plot(kullback_leibler~s1, data = divS1, type = "b", pch = 16,
     xlab = substitute(Sigma[1][1]), ylab = "Divergence")
axis(side = 1, at = Sigma2[1, 1], labels = substitute(Sigma[1][1]), cex.axis = 1.5)
```

#### Among the scale matrix: second situation

Let $\mathbf{X}^1 \sim MTD(\mathbf{0}, \nu, \boldsymbol{\Sigma}_1)$
and $\mathbf{X}^2 \sim MTD(\mathbf{0}, \nu, \boldsymbol{\Sigma}_2)$

where $\nu = 1$,  
$\boldsymbol{\Sigma}_2 = \left(\begin{array}{cc}
1 & 0.3 \\
0.3 & 1
\end{array}\right)$  
and $\boldsymbol{\Sigma}_1 = s \ \boldsymbol{\Sigma}_2$, $s \in ]0 ; 70]$.

The results are showed in @fig-divsigma12.

```{r}
#| cache: true

# The parameters
nu <- 1
Sigma <- matrix(c(1, 0.3, 0.3, 1), nrow = 2)

# Values of the proportionality coefficient
s <- c(seq(0.05, 1, by = 0.05), seq(2, 70, by = 2))

# Renyi divergence, beta = 0.25
b <- 0.25
renyi_0_25 <- sapply(s, function(s) {
  diststudent(nu1 = nu, Sigma1 = s*Sigma, nu2 = nu, Sigma2 = Sigma,
              dist = "renyi", bet = b)
})

# Renyi divergence, beta = 1.5
b <- 1.5
renyi_1_5 <- sapply(s, function(s) {
  diststudent(nu1 = nu, Sigma1 = s*Sigma, nu2 = nu, Sigma2 = Sigma,
              dist = "renyi", bet = b)
})

# Battacharyya divergence
bhattacharyya <- sapply(s, function(s) {
  diststudent(nu1 = nu, Sigma1 = s*Sigma, nu2 = nu, Sigma2 = Sigma,
              dist = "bhattacharyya")
})

# Hellinger divergence
hellinger <- sapply(s, function(s) {
  diststudent(nu1 = nu, Sigma1 = s*Sigma, nu2 = nu, Sigma2 = Sigma,
              dist = "hellinger")
})

# Kullback-Leibler divergence
kullback_leibler <- sapply(s, function(s) {
  kld(Sigma1 = s*Sigma, Sigma2 = Sigma, nu1 = nu, nu2 = nu,
      distribution = "mtd")
})

# Table of the results
divSigma <- data.frame(
  facteur = s,
  renyi_0_25 = renyi_0_25,
  renyi_1_5 = renyi_1_5,
  bhattacharyya = bhattacharyya,
  hellinger = hellinger,
  kullback_leibler = kullback_leibler
)
```

<!-- $\mathbf{X}^1$ et $\mathbf{X}^2$ -->

```{r}
#| label: fig-divsigma12
#| fig-cap: "Divergence between the distributions $\\mathbf{X}^1$ et $\\mathbf{X}^2$"
#| fig-subcap:
#|    - "Renyi ($\\beta = 0.25$)"
#|    - "Renyi ($\\beta = 1.5$)"
#|    - ""
#|    - "Bhattacharyya"
#|    - "Hellinger"
#|    - "Kullback-Leibler"
#| layout-ncol: 3
#| layout-nrow: 2

plot(renyi_0_25~facteur, data = divSigma, type = "b", pch = 16,
     xlab = "s", ylab = "Divergence")
axis(side = 1, at = 1, labels = 1, cex.axis = 1.5)

plot(renyi_1_5~facteur, data = divSigma, type = "b", pch = 16,
     xlab = "s", ylab = "Divergence")
axis(side = 1, at = 1, labels = 1, cex.axis = 1.5)

plot(0, 0, type = "n", bty = "n", xaxt = "n", yaxt = "n", xlab = "", ylab = "")

plot(bhattacharyya~facteur, data = divSigma, type = "b", pch = 16,
     xlab = "s", ylab = "Divergence")
axis(side = 1, at = 1, labels = 1, cex.axis = 1.5)

plot(hellinger~facteur, data = divSigma, type = "b", pch = 16,
     xlab = "s", ylab = "Divergence")
axis(side = 1, at = 1, labels = 1, cex.axis = 1.5)

plot(kullback_leibler~facteur, data = divSigma, type = "b", pch = 16,
     xlab = "s", ylab = "Divergence")
axis(side = 1, at = 1, labels = 1, cex.axis = 1.5)
```

<!-- # Other multivariate distributions  -->

## Multivariate Cauchy distribution

Let $\mathbf{X}$ be a random vector of $\mathbb{R}^p$ from a
multivariate Cauchy distribution (MCD) with location vector
$\boldsymbol{\mu}$ and scale matrix $\boldsymbol{\Sigma}$, which is a $p \times p$
symmetric, positive definite matrix. Its probability density function is
given by: $$
    \displaystyle{ f(\mathbf{x}|\boldsymbol{\mu}, \boldsymbol{\Sigma}) = \frac{\Gamma\left(\frac{1+p}{2}\right)}{\pi^{p/2} \Gamma\left(\frac{1}{2}\right) |\boldsymbol{\Sigma}|^\frac{1}{2} \left[ 1 + (\mathbf{x}-\boldsymbol{\mu})^T \boldsymbol{\Sigma}^{-1} (\mathbf{x}-\boldsymbol{\mu}) \right]^\frac{1+p}{2}} }
$$

where $\Gamma(\cdot)$ is the gamma function

The MCD is a particular case of the multivariate $t$-distribution with
$\nu = 1$. Hence, a sample from a MCD with location parameter
$\boldsymbol{\mu}$ and dispersion matrix $\boldsymbol{\Sigma}$ can be generated using
@eq-rmtd with $\nu = 1$:

$$
    \displaystyle{\mathbf{X} = \boldsymbol{\mu} + \frac{\mathbf{Y}}{\sqrt{U}}}
$$ {#eq-rmcd} where $\mathbf{Y}$ is generated from a centered Gaussian
distribution with covariance matrix $\boldsymbol{\Sigma}$ and $U$ is distributed
according to a Chi-squared distribution with $\nu$ degrees of freedom.

### Parameter estimation

To estimate the parameters of a MCD, we use the same method as for the
MTD (@sec-est_mtd) setting $\nu = 1$. Then the algorithm presented in
@sec-est_mtd becomes:

1.  Initialization of the parameters
    $\theta'=({\boldsymbol{\mu}}',{\boldsymbol{\Sigma}}')$

2.  **until** convergence **do**

    -   Compute
        $\mathbb{E}_{V|\mathbf{X}}\{V_i |\mathbf{x}_i,\theta'\}=\displaystyle{\frac{1+p}{1+(\mathbf{x}_i-\boldsymbol{\mu}')^T\boldsymbol{\Sigma}'^{-1}(\mathbf{x}_i-\boldsymbol{\mu}')}}$

    -   Compute
        $\mathbb{E}_{V|\mathbf{X}}\{\ln V_i|\mathbf{x}_i,\theta'\}=\psi(\frac{1+p}{2}) -\ln \frac{1}{2}\bigg[1+(\mathbf{x}_i-\boldsymbol{\mu}')^T\boldsymbol{\Sigma}'^{-1}(\mathbf{x}_i-\boldsymbol{\mu}')\bigg]$

    -   Estimate
        $\hat{\boldsymbol{\mu}} = \frac{\sum_{i=1}^{N}\mathbf{x}_i\mathbb{E}_{V|\mathbf{X}}\{V_i|\mathbf{x}_i,\theta'\}}{\sum_{i=1}^{N}\mathbb{E}_{V|\mathbf{X}}\{V_i|\mathbf{x}_i,\theta'\}}$

    -   Estimate
        $\hat{\boldsymbol{\Sigma}}=\frac{1}{N}\sum_{i=1}^{N}\mathbb{E}_{V|\mathbf{X}}\{V_i|\mathbf{x}_i,\theta'\}(\mathbf{x}_i-\boldsymbol{\mu}')(\mathbf{x}_i-\boldsymbol{\mu}')^T$

### Kullback-Leibler divergence

Let $\mathbf{X}^1$ and $\mathbf{X}^2$ be two random vectors of
$\mathbb{R}^p$ from MCDs with probability density functions:
$$\displaystyle{ f_{\mathbf{X}^1}\left(\mathbf{x}|\boldsymbol{\Sigma}_1\right) = f_{\mathbf{X}^1}\left(\mathbf{x}|\mathbf{0}, \boldsymbol{\Sigma}_1\right) = \frac{\Gamma\left(\frac{1+p}{2}\right)}{\pi^{p/2} \Gamma\left(\frac{1}{2}\right) |\boldsymbol{\Sigma}_1|^\frac{1}{2} \left[ 1 + \mathbf{x}^T \boldsymbol{\Sigma}_1^{-1} \mathbf{x} \right]^\frac{1+p}{2}} }$$
and
$$\displaystyle{ f_{\mathbf{X}^2}\left(\mathbf{x}|\boldsymbol{\Sigma}_2\right) = f_{\mathbf{X}^2}\left(\mathbf{x}|\mathbf{0}, \boldsymbol{\Sigma}_2\right) =  \frac{\Gamma\left(\frac{1+p}{2}\right)}{\pi^{p/2} \Gamma\left(\frac{1}{2}\right) |\boldsymbol{\Sigma}_2|^\frac{1}{2} \left[ 1 + \mathbf{x}^T \boldsymbol{\Sigma}_2^{-1} \mathbf{x} \right]^\frac{1+p}{2}} }$$
Let $\lambda_1, \dots, \lambda_p$ be the eigenvalues of the square
matrix $\boldsymbol{\Sigma}_1 \boldsymbol{\Sigma}_2^{-1}$, sorted in increasing order:
$\lambda_1 < \dots < \lambda_p$.

From @sec-kldt, the KLD to $\mathbf{X}^1$ from $\mathbf{X}^2$ (@eq-kl)
is given by (@bouhlel22):

$$
KL(X_1||X_2) = \displaystyle{-\frac{1}{2} \ln{ \prod_{i=1}^p{\lambda_i}} + \frac{1+p}{2} \times D}
$$

where $D$ depends on the values of $\lambda_1, \dots, \lambda_p$:

-   If $\lambda_1 < 1$ and $\lambda_p > 1$: $$
    \begin{aligned}
    D = & \displaystyle{ \bigg( \ln{\lambda_p} - } \\
    & \displaystyle{ \frac{\partial}{\partial a} \bigg\{ F_D^{(p)} \bigg( a, \underbrace{\frac{1}{2}, \dots, \frac{1}{2}, a + \frac{1}{2}}_p ; a + \frac{1+p}{2} ; 1 - \frac{\lambda_1}{\lambda_p}, \dots, 1 - \frac{\lambda_{p-1}}{\lambda_p}, 1 - \frac{1}{\lambda_p} \bigg) \bigg\}\bigg|_{a=0} \bigg) }
    \end{aligned}
    $$ {#eq-klmcd1}
-   If $\lambda_p < 1$: $$
      \displaystyle{ D = \frac{\partial}{\partial a} \bigg\{ F_D^{(p)} \bigg( a, \underbrace{\frac{1}{2}, \dots, \frac{1}{2}}_p ; a + \frac{1+p}{2} ; 1 - \lambda_1, \dots, 1 - \lambda_p \bigg) \bigg\}\bigg|_{a=0} }
      $$ {#eq-klmcd2}
-   If $\lambda_1 > 1$: $$
      \displaystyle{ D = \prod_{i=1}^p\frac{1}{\sqrt{\lambda_i}} \times \frac{\partial}{\partial a} \bigg\{ F_D^{(p)} \bigg( \frac{1+p}{2}, \underbrace{\frac{1}{2}, \dots, \frac{1}{2}}_p ; a + \frac{1+p}{2} ; 1 - \frac{1}{\lambda_1}, \dots, 1 - \frac{1}{\lambda_p} \bigg) \bigg\}\bigg|_{a=0} }
      $$ {#eq-klmcd3}

## Multivariate generalized Gaussian distribution

Let $\mathbf{X}$ a random vector of $\mathbb{R}^p$ from a Multivariate
generalized Gaussian distribution (MGGD) with mean vector
$\boldsymbol{\mu}$, dispersion matrix $\boldsymbol{\Sigma}$ (symmetric, positive
definite of dimension $p \times p$) and shape parameter $\beta > 0$. Its
probability density function is given by [@gomez; @bouhlel19]: $$
    \displaystyle{f_\mathbf{X}\left(\mathbf{x}|\boldsymbol{\mu}, \boldsymbol{\Sigma}, \beta\right) = \frac{\Gamma\left(\frac{p}{2}\right)}{\pi^\frac{p}{2} \Gamma\left(\frac{p}{2 \beta}\right) 2^\frac{p}{2\beta}} \frac{\beta}{|\boldsymbol{\Sigma}|^\frac{1}{2}} e^{-\frac{1}{2}\left[ (\mathbf{x}-\boldsymbol{\mu})^T \boldsymbol{\Sigma}^{-1} (\mathbf{x}-\boldsymbol{\mu}) \right]^\beta}}
$$ {#eq-mggd}

where $\Gamma(\cdot)$ is the gamma function.
<!-- Let $\mathbf{X}$ a random vector of $\mathbb{R}^p$ distributed according to a MGGD with mean vector $\boldsymbol{\mu} = \mathbf{0}$, dispersion matrix $\Sigma$ and shape parameter $\beta$. -->
@gomez show that $\mathbf{X}$ has the same distribution as: $$
    \displaystyle{\mathbf{X} = \tau \boldsymbol{\Sigma}^{1/2} \mathbf{U}}
$$ {#eq-rggd} where $\mathbf{U}$ is a random vector uniformly
distributed on the unit sphere in $\mathbb{R}^p$ and $\tau$ is such that
$\tau^{2\beta}$ is a random variable distributed according to a Gamma
distribution with shape parameter $\displaystyle{\frac{p}{2\beta}}$ and
scale parameter $2$. We use this property to simulate a sample from a
MGGD.

### Parameter estimation

Let $\left( \mathbf{x}_1, \dots, \mathbf{x}_n \right)$ be $n$ i.i.d.
random vectors distributed according to a MGGD with parameters
$\boldsymbol{\mu}$, $\boldsymbol{\Sigma}$ and $\beta$. @pascal give a method to
estimate these parameters.

The dispersion matrix $\boldsymbol{\Sigma}$ can be normalized as follows: $$
\boldsymbol{\Sigma} = m \mathbf M
$$ {#eq-normalizeSigma} $\mathbf M$ being a symmetric, positive-definite matrix
such that its trace is $tr(\mathbf M) = p$ and $m \in \mathbb{R}$.

The estimations of the parameters $(\boldsymbol{\mu}, \mathbf M, m, \beta)$
can be computed as follows:

-   The mean vector $\boldsymbol{\mu}$ is estimated by calculating the
    mean of $\left( \mathbf{x}_1, \dots, \mathbf{x}_n \right)$.
-   If $\beta$ is known, $\mathbf M$ and $m$ can be estimated by: $$
      \mathbf M = \frac{1}{n} \sum_{i=1}^n{\frac{n p}{\mathbf{y}_i + \mathbf{y}_i^{1-\beta} \sum_{j \neq i}{\mathbf{y}_j^\beta}} \mathbf{x}_i \mathbf{x}_i^T}
      $$ {#eq-estggdM} and $$
      m = \left( \frac{\beta}{p n} \sum_{i=1}^n{{\mathbf{y}_i}^\beta} \right)^\frac{1}{\beta}
      $$ {#eq-estggdm} with
    $\mathbf{y}_i = \mathbf{x}_i^T \mathbf M^{-1} \mathbf{x}_i$
-   When the shape parameter $\beta$ is unknown, it is estimated using the
    following relation: $$
      \displaystyle{\frac{p n}{2 \sum_{i=1}^n{\mathbf{y}_i^\beta}} \sum_{i=1}^n{\left( \mathbf{y}_i^\beta \ln{\mathbf{y}_i} \right)} - \frac{p n}{2\beta} \left( \Psi(\frac{p}{2\beta}) + \ln{2} \right) - n - \frac{p n}{2\beta} \ \ln{\left( \frac{\beta}{p n}\sum_{i=1}^n{\mathbf{y}_i^\beta} \right)} = 0}
      $$ {#eq-estggdbeta} where $\Psi$ is the digamma function:
    $\displaystyle{\Psi(x) = \frac{d}{dx} \ln{\Gamma(x)} = \frac{\Gamma'(x)}{\Gamma(x)}}$

Thus, an iterative algorithm is introduced by @pascal to estimate $\mathbf M$,
$m$ and $\beta$:

1.  Initialization of $\beta$ and $\mathbf M$

2.  **until** convergence **do**

    -   Estimation of $\mathbf M$ using one iteration of @eq-estggdM and
        normalization
    -   Estimation of $\beta$ using @eq-estggdbeta

    **end until**

3.  Estimation of $m$ using @eq-estggdm

### Kullback-Leibler divergence

Let $\mathbf{X}^1$ and $\mathbf{X}^2$ be two random vectors of
$\mathbb{R}^p$ ($p > 1$) from centered MGGDs with probability density
functions
$$\displaystyle{f_{\mathbf{X}^1}(\mathbf{x}|\boldsymbol{\Sigma}_1, \beta_1) = f_{\mathbf{X}^1}\left(\mathbf{x}|\mathbf{0}, \boldsymbol{\Sigma}_1, \beta_1\right) = \frac{\Gamma\left(\frac{p}{2}\right)}{\pi^\frac{p}{2} \Gamma\left(\frac{p}{2 \beta_1}\right) 2^\frac{p}{2\beta_1}} \frac{\beta_1}{|\boldsymbol{\Sigma}_1|^\frac{1}{2}} e^{-\frac{1}{2}\left[ \mathbf{x}^T \boldsymbol{\Sigma}_1^{-1} \mathbf{x} \right]^{\beta_1}}}$$
and
$$\displaystyle{f_{\mathbf{X}^2}(\mathbf{x}|\boldsymbol{\Sigma}_2, \beta_2) = f_{\mathbf{X}^2}\left(\mathbf{x}|\mathbf{0}, \boldsymbol{\Sigma}_2, \beta_2\right) = \frac{\Gamma\left(\frac{p}{2}\right)}{\pi^\frac{p}{2} \Gamma\left(\frac{p}{2 \beta_2}\right) 2^\frac{p}{2\beta_2}} \frac{\beta_2}{|\boldsymbol{\Sigma}_2|^\frac{1}{2}} e^{-\frac{1}{2}\left[ \mathbf{x}^T \boldsymbol{\Sigma}_2^{-1} \mathbf{x} \right]^{\beta_2}}}$$

The KLD of $\mathbf{X}^1$ from $\mathbf{X}^2$ is: $$
    \begin{array}{ll}
    \displaystyle{KL(\mathbf{X}^1||\mathbf{X}^2)} & \displaystyle{= \mathbb{E}_{X^1}\left[\ln\left(\frac{f_{\mathbf{X}^1}(\mathbf{X})}{f_{\mathbf{X}^2}(\mathbf{X})}\right)\right]} \\
    & \displaystyle{= \int_{\mathbb{R}^p}{\ln \left( \frac{f_{\mathbf{X}^1}(\mathbf{x}|\boldsymbol{\Sigma}_1, \beta_1)}{f_{\mathbf{X}^2}(\mathbf{x}|\boldsymbol{\Sigma}_2, \beta_2)} \right) f_{\mathbf{X}^1}(\mathbf{x}|\boldsymbol{\Sigma}_1, \beta_1)\mathrm{d}\mathbf{x}}}
    \end{array}
$$ {#eq-kl}

Let $\lambda_1, \dots, \lambda_p$ be the eigenvalues of the square
matrix $\boldsymbol{\Sigma}_1 \boldsymbol{\Sigma}_2^{-1}$ sorted in increasing order:
$\lambda_1 < \dots < \lambda_{p-1} < \lambda_p$. @bouhlel19 present an
analytical expression of the KLD to $\mathbf{X}^1$ from $\mathbf{X}^2$:

$$
    \begin{aligned}
    KL(\mathbf{X}^1||\mathbf{X}^2) = & \displaystyle{\ln\left( \frac{\beta_1 |\boldsymbol{\Sigma}_1|^{\frac{-1}{2}} \Gamma\left(\frac{p}{2\beta_2}\right)}{\beta_2 |\boldsymbol{\Sigma}_2|^{\frac{-1}{2}} \Gamma\left(\frac{p}{2\beta_1}\right)} \right) + \frac{p}{2}\left(\frac{1}{\beta_2} - \frac{1}{\beta_1}\right) \ln{2} - \frac{p}{2\beta_2} + 2^{\frac{\beta_2}{\beta_1}-1} \ \frac{\Gamma\left(\frac{\beta_2}{\beta_1}+\frac{p}{\beta_1}\right)}{\Gamma\left(\frac{p}{2\beta_1}\right)}\ {\lambda_p}^{\beta_2}} \\
    & \displaystyle{\times F_D^{(p-1)}\bigg(-\beta_2, \underbrace{\frac{1}{2}, \dots, \frac{1}{2}}_{p-1}, \frac{p}{2}; 1-\frac{\lambda_{p-1}}{\lambda_p}, \dots, 1-\frac{\lambda_1}{\lambda_p}\bigg)}
    \end{aligned}
$$ {#eq-klggd}

<!-- where $F_D^{(n)}$ is the Lauricella $D$-hypergeometric function defined -->
<!-- by: $$ -->
<!--     F_D^{(n)}\left(a; b_1, \dots, b_n; c; x_1, \dots, x_n\right) = \sum_{m_1=0}^\infty \dots \sum_{m_n=0}^\infty{\frac{(a)_{m_1+\dots+m_n} (b_1)_{m_1} \dots (b_n)_{m_n}}{(c)_{m_1+\dots+m_n}} \ \frac{{x_1}^{m_1}}{m_1!} \dots \frac{{x_n}^{m_n}}{m_n!}} -->
<!-- $$ {#eq-lauric} and the Pochhammer symbol $(q)_i$, where $q$ is a real -->
<!-- and $i$ is a non-negative integer, is given by: $$ -->
<!-- \displaystyle{ (q)_i = \frac{\Gamma(q+i)}{\Gamma(q)} =} \left\{ \begin{array}{ll} -->
<!-- 1 & \text{if } i = 0 \\ -->
<!-- \displaystyle{ q (q+1) \dots (q+i-1) } & \text{if } i \geq 1 -->
<!-- \end{array} \right. -->
<!-- $$ {#eq-pochhammer} -->

#### Univariate case : $p=1$ {.unnumbered .unlisted}
Let $X^1$ and $X^2$ be two random
variables distributed according to centered generalized Gaussian
distributions with probability density functions:
$$\displaystyle{f_{X^1}(x|\sigma_1, \beta_1) = f_{X^1}\left(x|0, \sigma_1, \beta_1\right) = \frac{\beta_1}{\Gamma\left(\frac{1}{2 \beta_1}\right) 2^\frac{1}{2\beta_1}\sqrt{\sigma_1}} e^{-\frac{1}{2}\left(\frac{x^2}{\sigma_1}\right)^\beta}}$$
and
$$\displaystyle{f_{X^2}(x|\sigma_2, \beta_2) = f_{X^2}\left(x|0, \sigma_2, \beta_2\right) = \frac{\beta_2}{\Gamma\left(\frac{2}{2 \beta_2}\right) 2^\frac{1}{2\beta_2}\sqrt{\sigma_2}} e^{-\frac{1}{2}\left(\frac{x^2}{\sigma_2}\right)^\beta}}$$

the KLD is given by: $$
    KL(X^1||X^2) = \displaystyle{\ln\left( \frac{\frac{\beta_1}{\sqrt{\sigma_1}} \Gamma\left(\frac{1}{2\beta_2}\right)}{\frac{\beta_2}{\sqrt{\sigma_2}} \Gamma\left(\frac{1}{2\beta_1}\right)} \right) + \frac{1}{2}\left(\frac{1}{\beta_2} - \frac{1}{\beta_1}\right) \ln{2} - \frac{1}{2\beta_2} + 2^{\frac{\beta_2}{\beta_1}-1} \ \frac{\Gamma\left(\frac{\beta_2}{\beta_1}+\frac{1}{\beta_1}\right)}{\Gamma\left(\frac{1}{2\beta_1}\right)}\ \left(\frac{\sigma_1}{\sigma_2}\right)^{\beta_2}}
$$

# The multvardiv package

<!-- ## Other probability distributions: Multivariate Cauchy distributions, Multivariate generalized Gaussian distributions -->

The [`multvardiv`](https://forge.inrae.fr/imhorphen/multvardiv) package
(@tbl-multvardiv) provides functions to compute the divergence between two
multivariate $t$ distributions. This section presents these functions,
with an example.
There are also functions to compute the divergence between two Cauchy or
generalized Gaussian distributions.

```{r}
#| label: "tbl-multvardiv"
#| echo: false
#| eval: true
#| output: asis
tabfcts <- data.frame(
  "Multivariate Distribution" = c("$t$", "Cauchy", "MGGD"),
  "Probability density function" = c("`dmtd`", "`dmc`", "`dmggd`"),
  "Simulation" = c("`rmtd`", "`rmc`", "`rmggd`"),
  "Parameter estimation" = c("`estparmtd`", "`estparmc`", "`estparmggd`"),
  "Rényi, Bhattacharyya or Hellinger divergence" = c("`dlstudent`", "", ""),
  "Kullback-Leibler divergence" = c("`kldstudent`", "`kldcauchy`", "`kldggd`")
)
knitr::kable(tabfcts,
             col.names = c("Multivariate Distribution", "Probability density function",
                           "Simulation", "Parameter estimation",
                           "Rényi, Bhattacharyya or Hellinger divergence",
                           "Kullback-Leibler divergence"),
             caption = "multvardiv package: the functions for MTD, MCD and MGGD")
```

Furthermore, two functions allow to plot the probability density of a
bivariate $t$, Cauchy or generalized Gaussian distribution:

-   `plotmvd`: 3D-plot
-   `contourmvd`: contour plot

Two other functions are useful for some computation:

*   `lauricella`: Lauricella D-Hypergeometric Function
<!-- *   `lnpochhammer`: logarithm of the Pochhammer Symbol -->
*   `pochhammer`: the Pochhammer symbol

## Example dataset: simulation from a multivariate $t$ distribution

As an example, we simulate two datasets.

```{r}
#| echo: true
#| include: false
#| eval: false
nu1 <- 2
Sigma1 <- rbind(c(2, 1.2, 0.4), c(1.2, 2, 0.6), c(0.4, 0.6, 2))
nu2 <- 4
Sigma2 <- rbind(c(1, 0.3, 0.1), c(0.3, 1, 0.4), c(0.1, 0.4, 1))
Sigma1
Sigma2
```

The `rmtd` function simulates a sample from a MTD:

```{r}
#| echo: true
set.seed(1234)
x1 <- rmtd(10000, nu = 2, mu = rep(0, 3), Sigma =
             rbind(c(2, 1.2, 0.4), c(1.2, 2, 0.6), c(0.4, 0.6, 2)))
head(x1)
```

```{r}
#| echo: true
set.seed(1234)
x2 <- rmtd(10000, nu = 4, mu = rep(0, 3), Sigma =
             rbind(c(1, 0.3, 0.1), c(0.3, 1, 0.4), c(0.1, 0.4, 1)))
head(x2)
```

## Parameter estimation

The `estparmtd` function allows to estimate the parameters of a MTD from
the data. It returns the estimations of the parameters, as a list of
three elements `$nu`, `$mu`, `$Sigma`:

```{r}
#| echo: true
par1 <- estparmtd(x1)
par1
```

```{r}
#| echo: true
par2 <- estparmtd(x2)
par2
```

As the `estparmtd` function uses an iterative algorithm, the returned
result has two attributes `attr(,"epsilon")` and `attr(,"k")`, which
indicate the precision of the estimation of $\nu$ and the number of
iterations that are necessary to reach this precision.

An optional argument `eps` (default $10^{-6}$) allows to give the
desired precision for the estimation of the `$nu` parameter:

```{r}
#| echo: true
estparmggd(x1, eps = 1e-4)
```

The precision `attr(,"epsilon")` is given by the `eps` argument.
<!-- except when the convergence was too slow and the precision could not be reached -->
<!-- (due to computational problems). -->
<!-- In the latter case, the value of `attr(,"epsilon")` is the precision that could actually be reached. -->

## Rényi, Bhattacharyya and Hellinger divergences

The `diststudentd` function computes the RD, BD or HD between two MTDs
with location parameter $\boldsymbol{\mu} = \mathbf{0}$.

By default, the `diststudent` function computes the RD of order $\beta$.
The order is given by the `bet` argument.

```{r}
#| echo: true
nu1 <- par1$nu; Sigma1 = par1$Sigma
nu2 <- par2$nu; Sigma2 = par2$Sigma
```

```{r}
#| echo: true
rd12 <- diststudent(nu1, Sigma1, nu2, Sigma2, bet = 0.5)
```

The calculation of the Lauricella $D$-hypergeometric function, involved
in the calculation of the RD between two MTDs (see @eq-rmtd) uses an
iterative algorithm. The optional `eps` argument (default `1e-6`) allows
to indicate the desired precision for this calculation:

```{r}
#| echo: true
#| eval: false
rd12 <- diststudent(nu1, Sigma1, nu2, Sigma2, bet = 0.5, eps = 1e-06)
```

The returned result is the RD with two attributes `attr(,"epsilon")`
(precision of the Lauricella function, given by the `eps` argument) and
`attr(,"k")` (number of iterations):

```{r}
#| echo: true
rd12
```

The optional `dist` argument allows to indicate the divergence:

-   Rényi (default):

    ```{r}
    #| echo: true
    diststudent(nu1, Sigma1, nu2, Sigma2, dist = "renyi", bet = 0.5)
    ```

-   Bhattacharyya:

    ```{r}
    #| echo: true
    diststudent(nu1, Sigma1, nu2, Sigma2, dist = "bhattacharyya")
    ```

-   Hellinger:

    ```{r}
    #| echo: true
    diststudent(nu1, Sigma1, nu2, Sigma2, dist = "hellinger")
    ```

The `bet` argument is required for the RD. It is omitted for the BD or
the HD.

## Kullback-Leibler divergence

The `kldstudent` function computes the KLD between two MTDs with
location parameter $\boldsymbol{\mu} = \mathbf{0}$.

We go on with the same example as for the Rényi divergence. The
`kldstudent` function computes the KLD:

```{r}
#| echo: true
kld12 <- kldstudent(nu1, Sigma1, nu2, Sigma2)
```

The calculation of the partial derivative of the Lauricella
$D$-hypergeometric function, needed to calculate the KLD between two
MTDs (see @eq-klmtd) uses an iterative algorithm. The optional `eps`
argument (default `1e-6`) allows to indicate the desired precision for
the calculation of this partial derivative:

```{r}
#| echo: true
#| eval: false
kld12 <- kldstudent(Sigma1, Sigma2, eps = 1e-06)
```

The returned result is the KLD with two attributes `attr(,"epsilon")`
(precision of the partial derivative, given by the `eps` argument) and
`attr(,"k")` (number of iterations):

```{r}
#| echo: true
kld12
```

## Execution time

### Among the wanted precision

How does the execution time increase when the precision increases?

We consider the distributions:
$X^1 \sim MTD(\mathbf{0}, \nu_1, \boldsymbol{\Sigma_1})$ and
$X^2 \sim MTD(\mathbf{0}, \nu_2, \boldsymbol{\Sigma_2})$

with $\nu_1 = 2$ and $\nu_2 = 4$,
$\boldsymbol{\Sigma_1} = \left(\begin{array}{ccc}
1 & 0.6 & 0.2 \\
0.6 & 1 & 0.3 \\
0.2 & 0.3 & 1
\end{array}\right)$
and
$\boldsymbol{\Sigma_2} = \left(\begin{array}{ccc}
1 & 0.3 & 0.1 \\
0.3 & 1 & 0.4 \\
0.1 & 0.4 & 1
\end{array}\right)$

```{r}
#| cache: true

nu1 <- 2
nu2 <- 4

times <- iter <- numeric(4)

Sigma1 = rbind(c(1, 0.6, 0.2), c(0.6, 1, 0.3), c(0.2, 0.3, 1))
Sigma2 = rbind(c(1, 0.3, 0.1), c(0.3, 1, 0.4), c(0.1, 0.4, 1))

epsilon <- c(1e-1, 1e-2, 1e-3, 1e-4, 1e-5, 1e-6, 1e-7, 1e-8, 1e-9, 1e-10, 1e-12,
             1e-13, 1e-14, 1e-15)
times <- iter <- numeric(length = length(epsilon))
for (i in 1:length(epsilon)) {
  eps <- epsilon[i]
  times[i] <- system.time(
    d <- diststudent(nu1 = nu1, Sigma1 = Sigma1, nu2 = nu2, Sigma2 = Sigma2,
                dist = "renyi", bet = 0.25, eps = eps)
  )["elapsed"]
  iter[i] <- attr(d, "k")
}["elapsed"]
```

```{r}
renyi_exetime <- data.frame(eps = epsilon, iter = iter, exetime = times)
```

```{r}
#| cache: true

epsilon <- c(1e-1, 1e-2, 1e-3, 1e-4, 1e-5, 1e-6, 1e-7, 1e-8, 1e-9, 1e-10, 1e-12,
             1e-13, 1e-14, 1e-15)
times <- iter <- numeric(length = length(epsilon))
for (i in 1:length(epsilon)) {
  eps <- epsilon[i]
  times[i] <- system.time(
    d <- kld(nu1 = nu1, Sigma1 = Sigma1, nu2 = nu2, Sigma2 = Sigma2,
           distribution = "mtd", eps = eps)
  )["elapsed"]
  iter[i] <- attr(d, "k")
}
```

```{r}
kl_exetime <- data.frame(eps = epsilon, iter = iter, exetime = times)
```

```{r}
library(kableExtra)
data.frame(renyi_exetime$eps, renyi_exetime[2:3], kl_exetime[2:3]) %>%
  kbl(col.names = c("Precision", "Number of iterations", "Execution time", "Number of iterations", "Execution time")) %>%
  add_header_above(c(" " = 1, "Rényi divergence" = 2, "Kullback-Leibler divergence" = 2))
```

### Among the dimension of the distributions

We fix the wanted precision: $\epsilon = 10^{-6}$.

```{r}
#| cache: true

nu1 <- 2
nu2 <- 4

times <- iter <- numeric(4)

# Dim 1
Sigma1 <- 0.2
Sigma2 <- 0.3
times[1] <- system.time(
  d <- diststudent(nu1 = nu1, Sigma1 = Sigma1, nu2 = nu2, Sigma2 = Sigma2,
              dist = "renyi", bet = 0.25)
)["elapsed"]
iter[1] <- attr(d, "k")

# Dim 2
Sigma1 = rbind(c(1, 0.2), c(0.2, 1))
Sigma2 = rbind(c(1, 0.3), c(0.3, 1))
times[2] <- system.time(
  d <- diststudent(nu1 = nu1, Sigma1 = Sigma1, nu2 = nu2, Sigma2 = Sigma2,
              dist = "renyi", bet = 0.25)
)["elapsed"]
iter[2] <- attr(d, "k")

# Dim 3
Sigma1 = rbind(c(1, 0.6, 0.2), c(0.6, 1, 0.3), c(0.2, 0.3, 1))
Sigma2 = rbind(c(1, 0.3, 0.1), c(0.3, 1, 0.4), c(0.1, 0.4, 1))
times[3] <- system.time(
  d <- diststudent(nu1 = nu1, Sigma1 = Sigma1, nu2 = nu2, Sigma2 = Sigma2,
              dist = "renyi", bet = 0.25)
)["elapsed"]
iter[3] <- attr(d, "k")

# Dim 4
Sigma1 = rbind(c(1, 0.6, 0.2, 0), c(0.6, 1, 0.3, 0),
               c(0.2, 0.3, 1, 0), c(0, 0, 0, 1))
Sigma2 = rbind(c(1, 0.3, 0.1, 0), c(0.3, 1, 0.4, 0),
               c(0.1, 0.4, 1, 0), c(0, 0, 0, 1))
times[4] <- system.time(
  d <- diststudent(nu1 = nu1, Sigma1 = Sigma1, nu2 = nu2, Sigma2 = Sigma2,
              dist = "renyi", bet = 0.25)
)["elapsed"]
iter[4] <- attr(d, "k")

# # Dim 5
# Sigma1 = rbind(c(1, 0.6, 0.2, 0, 0), c(0.6, 1, 0.3, 0, 0),
#                c(0.2, 0.3, 1, 0, 0), c(0, 0, 0, 1, 0),
#                c(0, 0, 0, 0, 1))
# Sigma2 = rbind(c(1, 0.3, 0.1, 0, 0), c(0.3, 1, 0.4, 0, 0),
#                c(0.1, 0.4, 1, 0, 0), c(0, 0, 0, 1, 0),
#                c(0, 0, 0, 0, 1))
# t5 <- system.time(
#   diststudent(nu1 = nu1, Sigma1 = Sigma1, nu2 = nu2, Sigma2 = Sigma2,
#               dist = "renyi", bet = 0.25)
# )
```

```{r}
renyi_exetime <- data.frame(dim = 1:4, iter = iter, exetime = times)
# knitr::kable(renyi_exetime,
#              col.names = c("Dimension p", "Number of iterations", "Execution time"),
#              caption = "Execution time among the dimension $p$ of the distributions, for the Rényi divergence with $\\beta = 0.25$.")
```

```{r}
#| cache: true

nu1 <- 2
nu2 <- 4

times <- iter <- numeric(4)

# Dim 1
Sigma1 <- 0.2
Sigma2 <- 0.3
times[1] <- system.time(
  d <- kld(nu1 = nu1, Sigma1 = Sigma1, nu2 = nu2, Sigma2 = Sigma2,
           distribution = "mtd")
)["elapsed"]
iter[1] <- attr(d, "k")

# Dim 2
Sigma1 = rbind(c(1, 0.2), c(0.2, 1))
Sigma2 = rbind(c(1, 0.3), c(0.3, 1))
times[2] <- system.time(
  d <- kld(nu1 = nu1, Sigma1 = Sigma1, nu2 = nu2, Sigma2 = Sigma2,
           distribution = "mtd")
)["elapsed"]
iter[2] <- attr(d, "k")

# Dim 3
Sigma1 = rbind(c(1, 0.6, 0.2), c(0.6, 1, 0.3), c(0.2, 0.3, 1))
Sigma2 = rbind(c(1, 0.3, 0.1), c(0.3, 1, 0.4), c(0.1, 0.4, 1))
times[3] <- system.time(
  d <- kld(nu1 = nu1, Sigma1 = Sigma1, nu2 = nu2, Sigma2 = Sigma2,
           distribution = "mtd")
)["elapsed"]
iter[3] <- attr(d, "k")

# Dim 4
Sigma1 = rbind(c(1, 0.6, 0.2, 0), c(0.6, 1, 0.3, 0),
               c(0.2, 0.3, 1, 0), c(0, 0, 0, 1))
Sigma2 = rbind(c(1, 0.3, 0.1, 0), c(0.3, 1, 0.4, 0),
               c(0.1, 0.4, 1, 0), c(0, 0, 0, 1))
times[4] <- system.time(
  d <- kld(nu1 = nu1, Sigma1 = Sigma1, nu2 = nu2, Sigma2 = Sigma2,
           distribution = "mtd")
)["elapsed"]
iter[4] <- attr(d, "k")

# # Dim 5
# Sigma1 = rbind(c(1, 0.6, 0.2, 0, 0), c(0.6, 1, 0.3, 0, 0),
#                c(0.2, 0.3, 1, 0, 0), c(0, 0, 0, 1, 0),
#                c(0, 0, 0, 0, 1))
# Sigma2 = rbind(c(1, 0.3, 0.1, 0, 0), c(0.3, 1, 0.4, 0, 0),
#                c(0.1, 0.4, 1, 0, 0), c(0, 0, 0, 1, 0),
#                c(0, 0, 0, 0, 1))
# t5 <- system.time(
#   diststudent(nu1 = nu1, Sigma1 = Sigma1, nu2 = nu2, Sigma2 = Sigma2,
#               dist = "renyi", bet = 0.25)
# )
```

```{r}
kl_exetime <- data.frame(dim = 1:4, iter = iter, exetime = times)
# knitr::kable(kl_exetime,
#              col.names = c("Dimension p", "Number of iterations", "Execution time"),
#              caption = "Execution time among the dimension $p$ of the distributions, for the Kullback-Leibler divergence.")
```

```{r}
library(kableExtra)
data.frame(renyi_exetime$dim, renyi_exetime[2:3], kl_exetime[2:3]) %>%
  kbl(col.names = c("Dimension", "Number of iterations", "Execution time", "Number of iterations", "Execution time")) %>%
  add_header_above(c(" " = 1, "Rényi divergence" = 2, "Kullback-Leibler divergence" = 2))
```

### Among the divergence

```{r}
#| cache: true

# Degrees of freedom
nu <- 2
Sigma2 <- matrix(c(1, 0.3, 0.1, 0.3, 1, 0.4, 0.1, 0.4, 1), nrow = 3)

# Values of the proportionality coefficient
S <- c(seq(0.1, 1, by = 0.1), seq(2, 70, by = 2))

# Rényi divergence
times <- iter <- numeric(length = length(S))
for (i in 1:length(S)) {
  s <- S[i]
  times[i] <- system.time(
    d <- diststudent(nu1 = nu, Sigma1 = s*Sigma, nu2 = nu, Sigma2 = Sigma,
                     dist = "renyi", bet = 0.25)
  )["elapsed"]
  iter[i] <- attr(d, "k")
}

```

```{r}
renyi_exetime <- data.frame(s = S, iter = iter, exetime = times)
plot(exetime~s, data = renyi_exetime, type = "b", pch = 16,
     xlab = "s", ylab = "Divergence")
axis(side = 1, at = 1, labels = 1, cex.axis = 1.5)
```

```{r}
#| cache: true

# Degrees of freedom
nu <- 2
Sigma2 <- matrix(c(1, 0.3, 0.1, 0.3, 1, 0.4, 0.1, 0.4, 1), nrow = 3)

# Values of the proportionality coefficient
S <- c(seq(0.1, 1, by = 0.1), seq(2, 50, by = 2))

# Kullback-Leibler divergence
times <- iter <- numeric(length = length(S))
for (i in 1:length(S)) {
  s <- S[i]
  times[i] <- system.time(
    d <- kld(nu1 = nu, Sigma1 = s*Sigma, nu2 = nu, Sigma2 = Sigma,
             distribution = "mtd")
  )["elapsed"]
  iter[i] <- attr(d, "k")
}
```

```{r}
kl_exetime <- data.frame(s = S, iter = iter, exetime = times)
plot(exetime~s, data = kl_exetime, type = "b", pch = 16,
     xlab = "s", ylab = "Execution time")
axis(side = 1, at = 1, labels = 1, cex.axis = 1.5)
```

## Other probability distributions

The [`multvardiv`](https://forge.inrae.fr/imhorphen/multvardiv) package
provides the same tools for the MCD and the MGGD (@tbl-multvardiv).

# Some particular cases -- test of the results

<!-- ## Package [`mggd`](https://CRAN.R-project.org/package=mggd), the `lauricella` function: some particular examples -->

## The `lauricella` function: some particular examples

The computation of the KLD between two MGGDs, as given by @eq-klggd,
includes the $D$-hypergeometric Lauricella function (@eq-lauric). For
some particular cases, a simpler calculation allows an exact computing
of this expression. We used them in the `tests` directory of the
[`multvardiv`](https://CRAN.R-project.org/package=multvardiv) package,
to test the `lauricella` function used by `kldggd` (see \[Divergence
between multivariate generalized Gaussian distributions\]).

### Case $p = 2$, $\beta_2=1$

When $\mathbf{X}^1$ and $\mathbf{X}^2$ are bivariate ($p = 2$), the
$D$-hypergeometric Lauricella function in @eq-klggd becomes: $$
\displaystyle{F_D^{(1)}\bigg(-\beta_2, \frac{1}{2}; 1; 1-\frac{\lambda_1}{\lambda_2}\bigg) = {_2F_1}\left(-\beta_2, \frac{1}{2}; 1; 1 - \frac{\lambda_1}{\lambda_2}\right)}
$$ {#eq-lauric2}

where ${_2F_1}$ is the Gauss hypergeometric function: $$
\displaystyle{{_2F_1}(a, b; c; x) = \sum_{m=0}^{\infty}{\frac{(a)_m (b)_m}{(c)_m} \frac{x^m}{m!}}}
$$ {#eq-lauric2next}

So, here: $$
\displaystyle{{_2F_1}\bigg(-1, \frac{1}{2}; 1; 1-\frac{\lambda_1}{\lambda_2}\bigg) = \frac{1}{2} \left(1 + \frac{\lambda_1}{\lambda_2}\right)}
$$ {#eq-lauric2end}

We use this result to test the result returned by the `lauricella`
function in this particular situation. Let us fix $p=2$, $\beta_2 = 1$,
$\lambda_1 = 1$, $\lambda_2 = 2$:

```{r}
#| echo: true
p <- 2
beta2 <- 1
lambda <- c(1, 2)
```

Let us compute $F_D^{(p-1)}$ using the `lauricella` function. It
returns:

```{r}
#| echo: true
lauricella2 <- lauricella(-beta2, 0.5, p/2, 1 - lambda[1]/lambda[2], eps = 1e-15)
lauricella2
```

and the value of the expression @eq-lauric2end is:

```{r}
#| echo: true
lauric2control <- 1/2*(1 + lambda[1]/lambda[2])
lauric2control
```

which is equal to `lauricella2`:

```{r}
#| echo: true
lauricella2 == lauric2control
```

### Case $p = 3$, $\beta_2=1$, $\lambda_1 = \lambda_2$

When $p = 3$, if $\beta_1 = 1$ and $\lambda_1 = \lambda_2$: If we note:
$\lambda = \lambda_1 = \lambda_2$, the $D$-hypergeometric Lauricella
function in @eq-klggd becomes: $$
\displaystyle{F_D^{(2)}\bigg(-1, \frac{1}{2}, \frac{1}{2}; \frac{3}{2}; 1-\frac{\lambda_1}{\lambda_3}, 1-\frac{\lambda_2}{\lambda_3}\bigg) = {_2F_1}\left(-1, 1; \frac{3}{2}; 1 - \frac{\lambda}{\lambda_3}\right) = \frac{1}{3} \left( 1 + \frac{2\lambda}{\lambda_3} \right)}
$$ {#eq-lauric3}

Let $p=3$, $\beta_2 = 1$, $\lambda_1 = \lambda_2 = 1$, $\lambda_3 = 2$:

```{r}
#| echo: true
p <- 3
beta2 <- 1
lambda <- c(1, 1, 2)
```

The calculation of $F_D^{(p-1)}$ using the `lauricella` function gives:

```{r}
#| echo: true
lauricella3 <- lauricella(-beta2, rep(0.5, p-1), p/2, 1 - lambda[1:2]/lambda[3],
                          eps = 1e-15)
lauricella3
```

and the value of the expression given by @eq-lauric3 is:

```{r}
#| echo: true
lauric3control <-1/3*(1 + 2*lambda[1]/lambda[3])
lauric3control
```

The results are equal. To be more precise, their difference is less than
$10^{-15}$:

```{r}
#| echo: true
as.numeric(lauricella3) - lauric3control
```

### Case $p = 4$, $\beta_2=1$, $\lambda_1 = \lambda_2 = \lambda_3$

We note : $\lambda = \lambda_1 = \lambda_2 = \lambda_3$.

In this case, the $D$-hypergeometric Lauricella function in @eq-klggd
becomes: $$
\begin{array}{ll}
\displaystyle{F_D^{(3)}\bigg(-1, \frac{1}{2}, \frac{1}{2}, \frac{1}{2}; 2; 1-\frac{\lambda_1}{\lambda_4}, 1-\frac{\lambda_2}{\lambda_4}, 1-\frac{\lambda_3}{\lambda_4}\bigg)} & = \displaystyle{{_2F_1}\left(-1, 3/2; 2; 1 - \frac{\lambda}{\lambda_p}\right)} \\
& = \displaystyle{\frac{1}{4} \left( 1 + \frac{3\lambda}{\lambda_p} \right)}
\end{array}
$$ {#eq-lauric4}

Let $p=4$, $\beta_2 = 1$, $\lambda_1 = \lambda_2 = \lambda_3 = 1$,
$\lambda_4 = 2$:

```{r}
#| echo: true
p <- 4
beta2 <- 1
lambda <- c(1, 1, 1, 2)
```

The calculation of $F_D^{(p-1)}$ using the `lauricella` function gives:

```{r}
#| echo: true
lauricella4 <- lauricella(-beta2, rep(0.5, p-1), p/2, 1 - lambda[1:3]/lambda[4],
                          eps = 1e-15)
lauricella4
```

and the value of the expression given by @eq-lauric4 is:

```{r}
#| echo: true
lauric4control <-1/4*(1 + 3*lambda[1]/lambda[4])
lauric4control
```

The results are equal:

```{r}
#| echo: true
lauricella4 == lauric4control
```

<!-- <!-- ## Package [`mcauchyd`](https://CRAN.R-project.org/package=mcauchyd), the `kldcauchy` function: some particular examples -->

--\> <!-- ## The `kldcauchy` function: some particular examples -->

<!-- For some particular cases, @bouhlel22 show a simpler calculation of the KLD between two MCDs. -->

<!-- We use the particular cases below to test the `kldcauchy` function presented in the section [Divergence between multivariate Cauchy distributions]. -->

<!-- ### Case $p = 1$ -->

<!-- When $X^1$ and $X^2$ are univariate, the KLD is [@bouhlel22]: -->

<!-- $$ -->

<!-- \displaystyle{KL(X^1||X^2) = \ln{\frac{\left( 1 + \sqrt{\lambda} \right)^2}{4 \sqrt{\lambda}} = KL(X^2||X^1)}} -->

<!-- $$ {#eq-klcd1} -->

<!-- Let $X^1$ be a random variable distributed according to a Cauchy distribution -->

<!-- with parameters $\mu_1 = 0$ and $\Sigma_1 = \frac{1}{2}$ -->

<!-- and $X^2$ a random variable distributed according to a Cauchy distribution -->

<!-- with parameters $\mu_2 = 0$ and $\Sigma_2 = 1$: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- Sigma1 <- 0.5 -->

<!-- Sigma2 <- 1 -->

<!-- ``` -->

<!-- Let us compute the KLD to $X^1$ from $X^2$, using `kldcauchy`: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- kl1_12 <- kldcauchy(Sigma1, Sigma2, eps = 1e-15) -->

<!-- kl1_12 -->

<!-- ``` -->

<!-- and the KLD to $X^2$ from $X^1$: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- kl1_21 <- kldcauchy(Sigma2, Sigma1, eps = 1e-15) -->

<!-- kl1_21 -->

<!-- ``` -->

<!-- Here, the $\lambda$ parameter is: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- lambda <- Sigma1/Sigma2 -->

<!-- lambda -->

<!-- ``` -->

<!-- The value of the expression given by @eq-klcd1 is: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- kl1control <- log( (1 + sqrt(lambda))^2 / (4*sqrt(lambda)) ) -->

<!-- kl1control -->

<!-- ``` -->

<!-- which is equal (to the precision $10^{-15}$) to `kl1_12` and `kl1_21`: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- as.numeric(kl1_12 - kl1control) -->

<!-- as.numeric(kl1_21 - kl1control) -->

<!-- ``` -->

<!-- ### Case $p = 2$, $\lambda_1 = \lambda_2$ -->

<!-- When $p=2$, if the eigenvalues of the matrix $\Sigma_1 \Sigma_2^{-1}$ are equal: -->

<!-- $\lambda_1 = \lambda_2 = \lambda$, the KLD is given by @bouhlel22: -->

<!-- $$ -->

<!-- \displaystyle{KL(\mathbf{X}^1||\mathbf{X}^2) = -\ln{\lambda} + \frac{3}{\sqrt{1 - \frac{1}{\lambda}}} \ln{\left(\sqrt{\lambda} + \sqrt{\lambda - 1}\right) - 3}} -->

<!-- $$ {#eq-kl12cd2} -->

<!-- and: -->

<!-- $$ -->

<!-- \displaystyle{KL(\mathbf{X}^2||\mathbf{X}^1) = \ln{\lambda} + \frac{3}{\sqrt{1 - \lambda}} \ln{\left(\sqrt{\frac{1}{\lambda}} + \sqrt{\frac{1}{\lambda} - 1}\right) - 3}} -->

<!-- $$ {#eq-kl21cd2} -->

<!-- Let $\mathbf{X}^1$ and $\mathbf{X}^2$ be two random vectors distributed according to Cauchy distributions -->

<!-- with parameters $(\boldsymbol{\mu}_1, \Sigma_1)$ and $(\boldsymbol{\mu}_2, \Sigma_2)$ -->

<!-- such that $\boldsymbol{\mu}_1 = \boldsymbol{\mu}_2 = \mathbf{0}$ and $\Sigma_1$ and $\Sigma_2$ are given by: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- Sigma1 <- diag(0.5, nrow = 2) -->

<!-- Sigma2 <- diag(1, nrow = 2) -->

<!-- Sigma1 -->

<!-- Sigma2 -->

<!-- ``` -->

<!-- Let us compute the KLD to $X^1$ from $X^2$, using `kldcauchy`: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- kl2_12 <- kldcauchy(Sigma1, Sigma2, eps = 1e-15) -->

<!-- kl2_12 -->

<!-- ``` -->

<!-- Given the eigenvalues of $\Sigma_1 \Sigma_2^{-1}$: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- lambda <- eigen(Sigma1 %*% solve(Sigma2), only.values = TRUE)$value -->

<!-- lambda -->

<!-- lambda <- unique(lambda) -->

<!-- ``` -->

<!-- The value of @eq-kl12cd2 is: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- lambda <- as.complex(lambda) -->

<!-- kl2control12 <- Re(-log(lambda) + -->

<!--                      3/sqrt(1-1/lambda)*log(sqrt(lambda) + sqrt(lambda-1)) - 3) -->

<!-- kl2control12 -->

<!-- ``` -->

<!-- which is equal (to the precision $10^{-15}$) to `kl2_12`: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- as.numeric(kl2_12 - kl2control12) -->

<!-- ``` -->

<!-- And the KLD to $X^2$ from $X^1$ is: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- kl2_21 <- kldcauchy(Sigma2, Sigma1, eps = 1e-15) -->

<!-- kl2_21 -->

<!-- ``` -->

<!-- and the value of the expression @eq-kl21cd2 is: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- kl2control21 <- Re(log(lambda) + -->

<!--                      3/sqrt(1-lambda)*log(sqrt(1/lambda) + sqrt(1/lambda-1)) - 3) -->

<!-- kl2control21 -->

<!-- ``` -->

<!-- which is equal (to the precision `1e-14`) to `kl2_21`: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- as.numeric(kl2_21 - kl2control21) -->

<!-- ``` -->

<!-- ### Case $p = 2$, $\lambda_1 \neq 1$, $\lambda_2 = 1$ -->

<!-- When $p = 2$ and either $\lambda_1 = 1$ and $\lambda_2 \neq 1$, or $\lambda_1 \neq 1$ and $\lambda_2 = 1$, @bouhlel22 show that if we note $\lambda_i$ the eigenvalue such that $\lambda_i \neq 1$: -->

<!-- $$ -->

<!-- KL(\mathbf{X}_1||\mathbf{X}_2) = \ln{\lambda_i} - \frac{3}{2} \frac{1}{\sqrt{1 - \lambda_i}} \ln{\left(\frac{1 - \sqrt{1 - \lambda_i}}{1 + \sqrt{1 - \lambda_i}}\right)} -->

<!-- $$ {#eq-klcd2} -->

<!-- Let $\mathbf{X}^1$ and $\mathbf{X}^2$ be two random vectors distributed according to Cauchy distributions -->

<!-- with parameters $(\boldsymbol{\mu}_1, \Sigma_1)$ and $(\boldsymbol{\mu}_2, \Sigma_2)$ -->

<!-- such that $\boldsymbol{\mu}_1 = \boldsymbol{\mu}_2 = \mathbf{0}$ and $\Sigma_1$ and $\Sigma_2$ are given by: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- Sigma1 <- matrix(c(0.5, 0, 0, 1), nrow = 2) -->

<!-- Sigma2 <- diag(nrow = 2) -->

<!-- Sigma1 -->

<!-- Sigma2 -->

<!-- ``` -->

<!-- Let us compute the KLD to $X^1$ from $X^2$, using `kldcauchy`: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- kl2_ex2 <- kldcauchy(Sigma1, Sigma2, eps = 1e-17) -->

<!-- kl2_ex2 -->

<!-- ``` -->

<!-- The eigenvalues of $\Sigma_1 \Sigma_2^{-1}$, sorted in increasing order, are: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- lambda <- sort(eigen(Sigma1 %*% solve(Sigma2), only.values = TRUE)$value) -->

<!-- lambda -->

<!-- lambdai <- lambda[1] -->

<!-- ``` -->

<!-- and the value of the expression @eq-klcd2 is: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- kl2control_ex2 <- log(lambdai) - -->

<!--   3/2 * 1/sqrt(1-lambdai)*log((1 - sqrt(1-lambdai))/(1 + sqrt(1-lambdai))) - 3 -->

<!-- kl2control_ex2 -->

<!-- ``` -->

<!-- which is equal (to the precision $10^{-15}$) to `kl2_ex2`: -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- as.numeric(kl2_ex2 - kl2control_ex2) -->

<!-- ``` -->

## The `kldstudent` function: some particular examples

Some particular cases allow a simpler calculation of the KLD between two
MTDs. we use these particular cases, showed hereafter, for some tests of
the `kldstudent` function.

### Case $p = 3$, $\lambda_1 = \lambda_2 = \lambda_3$

When $p = 3$, if the eigenvalues of the matrix $\Sigma_1 \Sigma_2^{-1}$
are equal: $\lambda_1 = \lambda_2 = \lambda_3 = \lambda$, if
$\lambda < 1$, we can get a simpler expression of the partial derivative
in @eq-klmt2.

Let: $$
f(\lambda) = \displaystyle{ \frac{\partial}{\partial{a}}{ \bigg\{ F_D^{(p)}\bigg( a, \frac{1}{2}, \frac{1}{2}, \frac{1}{2}; a + \frac{\nu_1+p}{2}; 1 - \frac{\nu_1}{\nu_2}\lambda_1, 1 - \frac{\nu_1}{\nu_2}\lambda_2, 1 - \frac{\nu_1}{\nu_2}\lambda_3 \bigg) \bigg\} } \bigg|_{a=0} }
$$ {#eq-kldmtdcase3}

$$
\begin{array}{lll}
f(\lambda) &    = &   \displaystyle{ \frac{\partial}{\partial a}\left\lbrace
_2 F_1 (a, \frac{3}{2},a+\frac{\nu_1+3}{2}, 1-\frac{\nu_1}{\nu_2}\lambda\right\rbrace\bigg|_{a=0} } \\
&   = &   \displaystyle{ \sum\limits_{n=0}^\infty \frac{\partial}{\partial a}\left\lbrace \frac{(a)_n}{(a+\frac{\nu_1+3}{2})_n}\right\rbrace\bigg|_{a=0} \bigg(\frac{3}{2}\bigg)_n\frac{(1-\frac{\nu_1}{\nu_2}\lambda)^n}{n!} } \\
&   = &   \displaystyle{ \sum\limits_{n=1}^\infty \frac{(1)_n}{(\frac{\nu_1+3}{2})_n} \frac{1}{n} \bigg(\frac{3}{2}\bigg)_n\frac{(1-\frac{\nu_1}{\nu_2}\lambda)^n}{n!} }
\end{array}
$$

The multiplication of the derivative of $f$ with respect to $\lambda$ by
$(1 -\frac{\nu_1}{\nu_2}\lambda)$ is given as follows:

$$
\begin{array}{lll}
(1-\frac{\nu_1}{\nu_2}\lambda)\frac{\partial}{\partial \lambda} f(\lambda) &    = &    -\frac{\nu_1}{\nu_2}\sum\limits_{n=1}^\infty \frac{(1)_n}{(\frac{\nu_1+3}{2})_n} \bigg(\frac{3}{2}\bigg)_n\frac{(1-\frac{\nu_1}{\nu_2}\lambda)^n}{n!}\\
&   = &   -\frac{\nu_1}{\nu_2}\left[ _2 F_1(1, \frac{3}{2}, \frac{\nu_1+3}{2}, 1-\frac{\nu_1}{\nu_2}\lambda) -1  \right]
\end{array}
$$

If $\nu_1=1$, Knowing that

$$_2 F_1(1, \frac{3}{2}, 2, 1-\frac{1}{\nu_2}\lambda) = \frac{2\nu_2}{\lambda+ \sqrt{\nu_2\lambda}}$$

we can state that:

$$\displaystyle{
\frac{\partial}{\partial \lambda} f(\lambda) = - 2\frac{\nu_2}{(\lambda+ \sqrt{\nu_2\lambda})(\nu_2-\lambda)} +\frac{1}{\nu_2-\lambda}
}$$

Accordingly,

$$\begin{array}{lll}
f(\lambda)    & = &   \displaystyle{ -\ln \frac{(\nu_2-\lambda)(\nu_2+\sqrt{\nu_2\lambda})}{\sqrt{\nu_2\lambda}-\nu_2}+2\frac{\sqrt{\nu_2}}{\sqrt{\nu_2}+\sqrt{\lambda}} + K }   \\
&   = &   \displaystyle{ -\ln \frac{\nu_2(1-\frac{\lambda}{\nu_2})(1+\sqrt{\frac{\lambda}{\nu_2}})}{\sqrt{\frac{\lambda}{\nu_2}}-1}+2\frac{\sqrt{\nu_2}}{\sqrt{\nu_2}+\sqrt{\lambda}} + K }    \\
&   = &   \displaystyle{ -\ln - \nu_2 \left(1 + \sqrt{\frac{\lambda}{\nu_2}} \right)^2 + 2 \frac{\sqrt{\nu_2}}{\sqrt{\nu_2} + \sqrt{\lambda}} + K }
\end{array}$$

The constant $K$ can be determined by putting
$f(\lambda=\frac{\nu_2}{\nu_1}) = 0$:

$$
-\ln\bigg[-\nu_2\left(1+\frac{1}{\sqrt{\nu_1}}\right)^2\bigg]+\frac{2}{1+\frac{1}{\sqrt{\nu_1}}} + K = 0
$$

Then

$$
K = \ln\bigg[-4\nu_2 \bigg]-1
$$

Accordingly,

$$
f(\lambda) = -2\ln \frac{(\nu_2+\sqrt{\nu_2\lambda})}{2\nu_2} + \frac{\sqrt{\nu_2}-\sqrt{\lambda}}{\sqrt{\nu_2}+\sqrt{\lambda}}
$$ {#eq-kldmtdcase3exact}

Let $\mathbf{X}^1$ and $\mathbf{X}^2$ be two random vectors distributed
according to $t$ distributions with parameters
$(\nu_1, \boldsymbol{\mu}_1, \boldsymbol{\Sigma}_1)$ and
$(\nu_2, \boldsymbol{\mu}_2, \boldsymbol{\Sigma}_2)$ such that
$\boldsymbol{\mu}_1 = \boldsymbol{\mu}_2 = 0$ and $\nu_1$, $\boldsymbol{\Sigma}_1$,
$\nu_2$ and $\boldsymbol{\Sigma}_2$ are given by:

```{r}
nu1 <- 1
Sigma1 <- diag(0.5, nrow = 3)
Sigma1
nu2 <- 2
Sigma2 <- diag(1, nrow = 3)
Sigma2
```

Let us compute the KLD to $\mathbf{X}^1$ from $\mathbf{X}^2$, using
`kldstudent`:

```{r}
kl3pc <- kldstudent(nu1, Sigma1, nu2, Sigma2, eps = .Machine$double.eps)
```

Given the eigenvalues of $\boldsymbol{\Sigma}_1 \boldsymbol{\Sigma}_2^{-1}$:

```{r}
lambda <- eigen(Sigma1 %*% solve(Sigma2), only.values = TRUE)$value
lambda
```

The value of @eq-kldmtdcase3exact is:

```{r}
lambda <- unique(lambda)
d <- -2*log((nu2 + sqrt(nu2*lambda))/(2*nu2)) + (sqrt(nu2) - sqrt(lambda))/(sqrt(nu2) + sqrt(lambda))
kl3control <- log( (gamma((nu1+3)/2) * gamma(nu2/2) * nu2^1.5) / (gamma((nu2+3)/2) * gamma(nu1/2) * nu1^1.5) ) +
  (nu2 - nu1)/2 * ( digamma((nu1+3)/2) - digamma(nu1/2) ) - 0.5*3*(log(lambda)) -
  (nu2 + 3)/2 * d
kl3control

```

Which is equal (to the precision `.Machine$double.eps`) to `kl3pc`:

```{r}
as.numeric(kl3pc) - kl3control
```

### Case $p = 4$, $\lambda_1 = \lambda_2 = \lambda_3 = \lambda_4$

Similarly, when $p = 4$, if the eigenvalues of the matrix
$\boldsymbol{\Sigma}_1 \boldsymbol{\Sigma}_2^{-1}$ are equal:
$\lambda_1 = \lambda_2 = \lambda_3 = \lambda_4 = \lambda$, if
$\lambda < 1$, we can get a simpler expression of the partial derivative
in @eq-klmt2.

Let: $$
f(\lambda) = \displaystyle{ \frac{\partial}{\partial{a}}{ \bigg\{ F_D^{(p)}\bigg( a, \frac{1}{2}, \frac{1}{2}, \frac{1}{2}, \frac{1}{2}; a + \frac{\nu_1+p}{2}; 1 - \frac{\nu_1}{\nu_2}\lambda_1, 1 - \frac{\nu_1}{\nu_2}\lambda_2, 1 - \frac{\nu_1}{\nu_2}\lambda_3, 1 - \frac{\nu_1}{\nu_2}\lambda_4 \bigg) \bigg\} } \bigg|_{a=0} }
$$ {#eq-kldmtdcase4}

$$\begin{array}{lll}
f(\lambda   & =   & \displaystyle{ \frac{\partial}{\partial a}\left\lbrace F_D^{(4)}\left(a, \frac{1}{2}, \frac{1}{2},\frac{1}{2},\frac{1}{2};a+\frac{\nu_1+4}{2}; 1-\frac{\nu_1}{\nu_2}\lambda,1-\frac{\nu_1}{\nu_2}\lambda,1-\frac{\nu_1}{\nu_2}, 1-\frac{\nu_1}{\nu_2}\lambda\right) \right\rbrace\bigg|_{a=0} }  \\
&   = &   \displaystyle{ \frac{\partial}{\partial a}\left\lbrace
_2 F_1 (a, 2,a+\frac{\nu_1+4}{2}, 1-\frac{\nu_1}{\nu_2}\lambda\right\rbrace\bigg|_{a=0} }    \\
&   = &   \displaystyle{ \sum\limits_{n=0}^\infty \frac{\partial}{\partial a}\left\lbrace \frac{(a)_n}{(a+\frac{\nu_1+4}{2})_n}\right\rbrace\bigg|_{a=0} (2)_n\frac{(1-\frac{\nu_1}{\nu_2}\lambda)^n}{n!} } \\
&   = &   \displaystyle{ \sum\limits_{n=1}^\infty \frac{(1)_n}{(\frac{\nu_1+4}{2})_n} \frac{1}{n} (2)_n\frac{(1-\frac{\nu_1}{\nu_2}\lambda)^n}{n!} }
\end{array}$$

The multiplication of the derivative of $f$ with respect to $\lambda$ by
$(1 -\frac{\nu_1}{\nu_2}\lambda)$ is given as follows:

$$\begin{array}{lll}
\displaystyle{ (1-\frac{\nu_1}{\nu_2}\lambda)\frac{\partial}{\partial \lambda} f(\lambda) } &   = &   \displaystyle{ -\frac{\nu_1}{\nu_2}\sum\limits_{n=1}^\infty \frac{(1)_n}{(\frac{\nu_1+4}{2})_n} (2)_n\frac{(1-\frac{\nu_1}{\nu_2}\lambda)^n}{n!} }    \\
&   = &   \displaystyle{ -\frac{\nu_1}{\nu_2}\left[ _2 F_1(1, 2, \frac{\nu_1+4}{2}, 1-\frac{\nu_1}{\nu_2}\lambda) -1  \right] }
\end{array}
$$

If $\nu_1=2$, Knowing that:

$$
\displaystyle{ _2 F_1(1, 2, 3, 1-\frac{2}{\nu_2}\lambda) = -\frac{2\nu_2(\nu_2-2\lambda+\nu_2\ln(\frac{2\lambda}{\nu_2}))}{(\nu_2-2\lambda)^2} }
$$

we can state that:

$$\begin{array}{lll}
\displaystyle{ \frac{\partial}{\partial \lambda} f(\lambda) } &    = &   \displaystyle{ -\frac{2}{\nu_2}\left[ -\frac{2\nu_2(\nu_2-2\lambda+\nu_2\ln(\frac{2\lambda}{\nu_2}))}{(\nu_2-2\lambda)^2(1-\frac{2}{\nu_2}\lambda)} -\frac{1}{1-\frac{2}{\nu_2}\lambda}  \right] }   \\
&   = &   \displaystyle{ \frac{4\nu_2(\nu_2-2\lambda+ \nu_2\ln(\frac{2\lambda}{\nu_2}))} {(\nu_2-2\lambda)^3} + \frac{2}{\nu_2-2\lambda} }
\end{array}$$

Accordingly,

$$
\displaystyle{ f(\lambda) = \frac{\nu_2}{\nu_2-2\lambda}-\ln(\lambda)+\frac{\nu_2^2\ln(\frac{2\lambda}{\nu_2})}{(\nu_2-2\lambda)^2} + K }
$$

The constant $K$, can be determined by using Taylor series expansions of
$f$:

$$
f(\lambda) = K -\frac{1}{2} - \ln(\frac{\nu_2}{2}) + \frac{4(\frac{\nu_2}{2} - \lambda)}{3\nu_2} + \frac{(\frac{\nu_2}{2} - \lambda)^2}{\nu_2^2} + \dots
$$

and then putting $f(\lambda = \frac{\nu_2}{2}) = 0$,

$$
K = \frac{1}{2}+\ln(\frac{\nu_2}{2})
$$

As a consequence:

$$
f(\lambda) = \frac{\nu_2}{\nu_2-2\lambda}-\ln(\frac{2\lambda}{\nu_2})+\frac{\nu_2^2\ln(\frac{2\lambda}{\nu_2})}{(\nu_2-2\lambda)^2} +\frac{1}{2}
$$

# Conclusion and perspectives

The [`multvardiv`](https://CRAN.R-project.org/package=multvardiv)
package provides tools for multivariate generalized Gaussian
distributions, multivariate $t$-distributions and multivariate Cauchy
distributions, the latter being a particular case of the $t$
distribution. Some of them are the Kullback-Leibler divergence and, for
the MTD and MCD, the Rényi, Bhattacharyya and Hellinger divergences,
based on analytical expressions given by @bouhlel19, @bouhlel22 and
@bouhlel23. Using some particular cases which allow an exact calculation
of the Lauricella function, we could test the `lauricella` function used
by the `kldggd` function of the
[`multvardiv`](https://CRAN.R-project.org/package=multvardiv) package.
Similarly, we used some particular cases for the calculation of the KLD
between MTDs or between MCDs to test the `kldtstudent` and `kldcauchy`
function<!-- of the [`mcauchyd`](https://CRAN.R-project.org/package=mcauchyd) package -->.

In a future version of the
[`multvardiv`](https://CRAN.R-project.org/package=multvardiv) package,
we will add other methods for the parameter estimation of a MGGD. In
further works, we will extend these tools to complex multivariate
distributions, and more generally to elliptical distributions.

<!-- # Conclusion and perspectives -->

<!-- # References {.unnumbered} -->

<!-- ::: {#refs} -->

<!-- ::: -->
